---
title: "Traffic Jams in Fish Size Spectra: Setting Up and Exploring the Model"
author: "Gustav Delius"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
  cache: true
---

## Background

The central hypothesis of this project is that fish population dynamics can
exhibit a **traffic jam instability** analogous to phantom jams on motorways.
In the McKendrick–von Foerster equation

$$\frac{\partial N(w)}{\partial t} = -\frac{\partial}{\partial w}\big[g(w)\,N(w)\big] - \mu(w)\,N(w),$$

the growth rate $g(w)$ plays the role of vehicle velocity. A traffic jam forms
when high fish density at size $w$ depletes the prey resource at size
$w' \approx w/\beta$ ($\beta \approx 100$ is the preferred predator–prey mass
ratio), which reduces $g(w)$, which causes fish to linger at $w$, further
depleting the resource — a self-reinforcing feedback.

The plankton resource follows semi-chemostat dynamics:
$$\frac{dN_R}{dt} = r_R\,(c_R - N_R) - \mu_R\,N_R,$$
giving equilibrium abundance $N_R^* = r_R\,c_R / (r_R + \mu_R)$.

When $r_R$ is small relative to $\mu_R$ (slow replenishment), the resource
level $N_R^*/c_R = r_R/(r_R + \mu_R)$ is low, predation strongly depletes the
resource, and any increase in fish density causes a large relative drop in
$N_R^*$:
$$\frac{\delta N_R^*}{N_R^*} = -\frac{\delta\mu_R}{r_R + \mu_R}.$$
This coupling between fish density and growth rate is maximised when
$r_R \ll \mu_R$.

## Model setup

```{r setup}
library(mizer)
library(ggplot2)
```

### `newSingleSpeciesParams()` and `setResource()`

We use `newSingleSpeciesParams()` to place a single species in a power-law
background and `setResource()` to switch the resource from the default
fixed background (`resource_constant`) to semi-chemostat dynamics
(`resource_semichemostat`) with a controllable replenishment rate.

The feeding level

$$f(w) = \frac{E(w)}{E(w) + h\,w^n}$$

is the fraction of maximum intake rate actually achieved. The growth rate

$$g(w) = \alpha\,h\,w^n\,[f(w) - f_c]$$

is zero when $f = f_c$ (the critical feeding level, default 0.25) and rises
linearly above it. Setting `f0 = 0.3` — just above $f_c = 0.25$ — places fish
in a regime where any reduction in resource abundance dramatically reduces
growth, exactly as required for the jam.

```{r params}
params_base <- newSingleSpeciesParams(
  species_name       = "Target species",
  w_max              = 1e4,     # 10 kg
  kappa              = 0.005,
  f0                 = 0.3,     # close to fc = 0.25
  reproduction_level = 0
)

# All three scenarios use r_R = 0.1 via setResource(); only f0 differs.
# Changing r_R with balance = TRUE does not change the jam dynamics because
# the equilibrium feeding level stays at f0. The role of r_R is indirect:
# in a real ecosystem a low r_R depletes the resource and pushes f0 toward fc.
params_low_r  <- setResource(params_base, resource_rate = 0.1,
                             resource_dynamics = "resource_semichemostat")
params_high_r <- setResource(params_base, resource_rate = 10,
                             resource_dynamics = "resource_semichemostat")

params_mid <- newSingleSpeciesParams(
  species_name       = "Target species",
  w_max              = 1e4,
  kappa              = 0.005,
  f0                 = 0.4,     # moderate distance from fc
  reproduction_level = 0
) |>
  setResource(resource_rate = 0.1, resource_dynamics = "resource_semichemostat")

params_stable <- newSingleSpeciesParams(
  species_name       = "Target species",
  w_max              = 1e4,
  kappa              = 0.005,
  f0                 = 0.6,     # well above fc; insensitive to prey changes
  reproduction_level = 0
) |>
  setResource(resource_rate = 0.1, resource_dynamics = "resource_semichemostat")
```

## What `balance = TRUE` does

`setResource()` uses `balance = TRUE` by default: it adjusts the carrying
capacity $c_R(w)$ so that the current resource abundance $N_R(w)$ is already
at its semi-chemostat equilibrium. As a result **both the low- and high-$r_R$
models start with the same feeding level** $f_0 = 0.3$.

```{r balance}
cat("Resource level N_R/c_R at t = 0 (equilibrium by balance = TRUE):\n")
cat("  low  r_R = 0.1:", round(range(resource_level(params_low_r)),  3), "\n")
cat("  high r_R = 10 :", round(range(resource_level(params_high_r)), 3), "\n")
```

The range `[0.33, 1]` for the low-$r_R$ case means that at the prey sizes of
the abundant small fish, $r_R \ll \mu_R$ and the resource is heavily depleted.
The `balance = TRUE` setting compensates by setting $c_R$ very high, preserving
the equilibrium. **The relevant implication for real ecosystems**: in a system
where $r_R$ is genuinely small, a naturally established fish population would
settle at a lower feeding level (nearer $f_c$) because the resource cannot
sustain a high $N_R^*$.

## The coupling mechanism

Growth rate collapses near $f_c$ when resource is even slightly depleted.

The key point is that `resource_level(p) <- rl` adjusts the *carrying
capacity* $c_R$, not the actual resource abundance $N_R$; `getEGrowth()`
uses $N_R$, so we must scale `initialNResource()` directly.

With $f_0 = 0.3$ and $f_c = 0.25$, the growth rate is zero when
$$N_R \leq N_R^{\,\text{base}} \times \frac{(1-f_0)\,f_c}{f_0(1-f_c)} \approx 0.778\,N_R^{\,\text{base}}.$$
So a mere 22% drop in resource abundance is enough to halt all growth — the
fish analogue of a motorway grinding to a complete standstill.

```{r coupling-plot}
#| fig-cap: "Growth rate g(w) when N_R is scaled to different fractions of its baseline value. With f₀ = 0.3 (only 20% above fc = 0.25), a 10% drop in N_R halves g(w); a 20% drop leaves only 11% of normal growth."
#| fig-width: 8
#| fig-height: 4.5

# Use depletion levels where g > 0; rl > (1-f0)*fc / (f0*(1-fc)) ~ 0.778
depletion_df <- do.call(rbind, lapply(c(1.0, 0.95, 0.90, 0.85, 0.80), function(rl) {
  p <- params_base
  initialNResource(p) <- initialNResource(p) * rl   # actually reduce N_R
  data.frame(
    resource_level = factor(rl),
    w              = w(p),
    g              = getEGrowth(p)["Target species", ]
  )
}))

ggplot(depletion_df, aes(x = w, y = g, colour = resource_level)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10(labels = scales::label_log(), breaks = 10^(-2:4)) +
  scale_colour_viridis_d(direction = -1,
                         name = expression(N[R] ~ "(fraction of baseline)")) +
  labs(x = "Body mass w (g)", y = "Growth rate g(w) (g/yr)") +
  theme_bw(base_size = 13)
```

This is the **traffic-jam mechanism**: fish growing through a size class where
the resource has been depleted find themselves nearly unable to grow. They
linger, consuming more resource, and the jam is self-reinforcing.

## Where is the coupling strongest?

Resource depletion (low $N_R/c_R$) occurs where predation mortality $\mu_R$ is
large relative to $r_R$. In the single-species model this is at the smallest
sizes, where the many small fish eat heavily.

```{r resource-level-plot}
#| fig-cap: "Resource level across the full size spectrum for two replenishment rates. Low r_R creates strong depletion where fish predation is heavy (small prey sizes). High r_R keeps the resource near capacity everywhere."
#| fig-width: 8
#| fig-height: 4.5

wf   <- w_full(params_low_r)
rl_df <- data.frame(
  w     = rep(wf, 2),
  level = c(resource_level(params_low_r), resource_level(params_high_r)),
  model = rep(c("low r_R = 0.1", "high r_R = 10"), each = length(wf))
)
rl_df <- rl_df[is.finite(rl_df$level) & rl_df$level > 0, ]

ggplot(rl_df, aes(x = w, y = level, colour = model)) +
  geom_line(linewidth = 1) +
  scale_x_log10(labels = scales::label_log(), breaks = 10^(-12:4)) +
  scale_colour_manual(values = c("firebrick", "steelblue"), name = NULL) +
  labs(x = "Resource size w (g)",
       y = expression(N[R]/c[R] ~ "(resource level)")) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

**Key finding**: with $r_R = 0.1$, the resource is depleted to ~33% of
capacity at the prey sizes of the smallest fish (where $\mu_R \approx 3 \gg r_R$).
With $r_R = 10$, the resource stays near capacity everywhere. In a real
ecosystem, the low-$r_R$ state would naturally correspond to near-$f_c$ feeding
for the fish living at those sizes.

## Demonstrating the traffic jam

### Perturbation experiment

We add a Gaussian density excess (4× normal density) centred at $w = 50\,\mathrm{g}$
and track how it evolves over 30 years in three scenarios:

| Scenario | $f_0$ | $r_R$ | Expected behaviour |
|---|---|---|---|
| Jam-prone | 0.3 | 0.1 | Slow wave; long-lived excess |
| Intermediate | 0.4 | 0.1 | Wave decays over ~15 yr |
| Stable | 0.6 | 0.1 | Fast wave; excess dissipates in ~10 yr |

All three use the same `r_R = 0.1` via `setResource()`. Varying $r_R$ with
`balance = TRUE` does **not** change the jam signal because the equilibrium
feeding level stays fixed at $f_0$. The role of a low $r_R$ is therefore
**indirect**: in a real ecosystem, depleted plankton (low $N_R^*/c_R$)
naturally pushes the fish feeding level toward $f_c$, entering the jam-prone
regime shown by the red curve here.

```{r perturbation-setup}
add_bump <- function(params, w_bump = 50, amplitude = 4, log10_sigma = 0.4) {
  N0   <- initialN(params)
  wv   <- w(params)
  bump <- amplitude * exp(-((log10(wv) - log10(w_bump))^2) / (2 * log10_sigma^2))
  N0[1, ] <- N0[1, ] * (1 + bump)
  initialN(params) <- N0
  params
}

run_pair <- function(params, t_max = 30) {
  base <- project(params, t_max = t_max, effort = 0, progress_bar = FALSE)
  pert <- project(add_bump(params), t_max = t_max, effort = 0, progress_bar = FALSE)
  list(base = base, pert = pert)
}

sims_jam    <- run_pair(params_low_r)
sims_mid    <- run_pair(params_mid)
sims_stable <- run_pair(params_stable)
```

```{r wave-extraction}
get_wave <- function(sim_base, sim_pert, label) {
  N_b   <- N(sim_base)[, 1, ]
  N_p   <- N(sim_pert)[, 1, ]
  times <- as.numeric(dimnames(N_b)[[1]])
  wv    <- w(sim_base@params)
  do.call(rbind, lapply(seq_along(times), function(i) {
    r  <- N_p[i, ] / N_b[i, ]
    ok <- is.finite(r) & N_b[i, ] > 0 & r > 0 & wv > 0.01
    if (!any(ok)) return(NULL)
    data.frame(time = times[i],
               amplitude = max(r[ok]),
               peak_w    = wv[ok][which.max(r[ok])],
               scenario  = label)
  }))
}

wave_df <- rbind(
  get_wave(sims_jam$base,    sims_jam$pert,    "jam-prone: f0=0.3 (near fc=0.25)"),
  get_wave(sims_mid$base,    sims_mid$pert,    "intermediate: f0=0.4"),
  get_wave(sims_stable$base, sims_stable$pert, "stable: f0=0.6")
)
wave_df$scenario <- factor(wave_df$scenario, levels = unique(wave_df$scenario))
```

### Jam amplitude

```{r amplitude-plot}
#| fig-cap: "Peak density excess over time (all three runs use r_R = 0.1). Proximity to fc = 0.25 is the key variable: f₀ = 0.3 keeps the jam alive for 30+ years; f₀ = 0.4 decays within 15 years; f₀ = 0.6 is gone by t = 10 yr."
#| fig-width: 8
#| fig-height: 4.5

ggplot(wave_df, aes(x = time, y = amplitude, colour = scenario)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("firebrick", "darkorange", "steelblue"),
                      name = NULL) +
  labs(x = "Time (years)",
       y = "Peak ratio  N_perturbed / N_baseline") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

### Wave speed

```{r speed-plot}
#| fig-cap: "Location of the density peak over time (log scale). With f₀ = 0.3 the peak advances from 50 g to only ~1000 g in 30 years (slow traffic). With f₀ = 0.4 it reaches ~5000 g; with f₀ = 0.6 it exits the domain within 10 years."
#| fig-width: 8
#| fig-height: 4.5

ggplot(wave_df, aes(x = time, y = peak_w, colour = scenario)) +
  geom_line(linewidth = 1) +
  scale_y_log10(labels = scales::label_log(), breaks = 10^(0:4)) +
  scale_colour_manual(values = c("firebrick", "darkorange", "steelblue"),
                      name = NULL) +
  labs(x = "Time (years)",
       y = "Body mass of density peak (g)") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

### Heatmap

```{r heatmap}
#| fig-cap: "Size–time heatmap of log₂(N_perturbed / N_baseline) for the jam-prone scenario. The red band advances slowly, confirming that fish near the critical feeding level propagate density excesses much more slowly than well-fed fish."
#| fig-width: 9
#| fig-height: 5.5

N_b <- N(sims_jam$base)[, 1, ]
N_p <- N(sims_jam$pert)[, 1, ]
wv  <- w(params_low_r)
tv  <- as.numeric(dimnames(N_b)[[1]])

ratio_mat <- N_p / N_b
ratio_mat[!is.finite(ratio_mat) | ratio_mat <= 0] <- NA
hm_df <- data.frame(
  w          = rep(wv,  each  = length(tv)),
  time       = rep(tv,  times = length(wv)),
  log2_ratio = as.vector(log2(ratio_mat))
)

ggplot(hm_df, aes(x = w, y = time, fill = log2_ratio)) +
  geom_raster() +
  scale_x_log10(labels = scales::label_log(), breaks = 10^(0:4)) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1.5, 1.5),
                       oob = scales::squish, na.value = "white",
                       name = expression(log[2](N/N[base]))) +
  labs(x = "Body mass w (g)", y = "Time (years)") +
  theme_bw(base_size = 13)
```

## Summary of findings

```{r summary-table, echo=FALSE}
#| tbl-cap: "Persistence of the density excess: amplitude at key time points."
library(knitr)
tab <- do.call(rbind, lapply(levels(wave_df$scenario), function(scen) {
  d <- wave_df[wave_df$scenario == scen, ]
  data.frame(
    Scenario  = scen,
    `t = 0`  = round(d$amplitude[d$time == 0],  2),
    `t = 10` = round(d$amplitude[d$time == 10], 2),
    `t = 30` = round(d$amplitude[d$time == 30], 2),
    `Decay`  = paste0(round(100*(1 - d$amplitude[d$time == 30] /
                                     d$amplitude[d$time == 0]), 0), "%"),
    check.names = FALSE
  )
}))
kable(tab, row.names = FALSE)
```

### Conclusions

1. **Near-critical feeding is the key condition.** With $f_0 = 0.3$ (only 20%
   above $f_c = 0.25$), the density excess decays by only 43% over 30 years.
   With $f_0 = 0.6$ it dissipates within 10 years. This is the fish analogue
   of a traffic queue: slow-moving vehicles preserve congestion far longer
   than fast-moving ones.

2. **Low $r_R$ creates the near-critical state in real ecosystems.** The
   resource is depleted to ~33% of capacity at the prey sizes of the abundant
   small fish when $r_R = 0.1$. A real fish population that self-regulates
   against this depleted resource would naturally settle at a feeding level
   near $f_c$, placing the system in the jam-prone regime.

3. **The density-dependent coupling (resource depletion by the bump fish) is
   weak in this single-species model.** The perturbation at $w = 50\,\mathrm{g}$
   causes negligible additional depletion at the preferred prey size
   ($w \approx 0.5\,\mathrm{g}$) because the predation mortality on the
   resource at that size is already tiny ($\mu_R \approx 10^{-7}$ yr$^{-1}$)
   relative to $r_R$. The jam signal shown here is primarily a **kinematic**
   effect (slow propagation near $f_c$), not yet a fully self-sustaining
   **dynamic** instability.

4. **A stronger coupling — and therefore a self-amplifying jam — is expected
   in a multispecies model** where fish eat other fish. In such a model, a
   dense cohort at size $w$ directly depletes the fish at size $w/\beta$,
   creating a strong non-local feedback that can push the system past the
   instability threshold.

### Next steps

- Set up a community or trait-based model (`newCommunityParams()` or
  `newTraitParams()`) with inter-species predation to get the stronger coupling.
- Search for the bifurcation parameter (perhaps $r_R$ itself, or fish density)
  at which the jam transitions from a decaying wave to a growing instability.
- Investigate whether targeted fishing at the jam size class can dissolve the
  jam and restore productivity — the central hypothesis of the project.
