---
title: "Day 21: Watching Hysteresis Play Out in Time, Not Just in the Bifurcation Diagram"
author: "Samia"
date: "2026-07-06"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, limit-cycle, hysteresis, time-series, diffusion, blog]
description: "Biomass is tracked over time (not just settled max/min), first independently, then as a chained forward/backward pair that shows clear hysteresis at resource_decrease = 0.05. The independent run also turns up something odd: higher resource_decrease values, where more of the default resource rate survives, are the ones that go extinct. A preliminary, backward-first bifurcation diagram and a couple of other unfinished threads are parked at the end."
---

## Where Day 20 Left Us

Day 20 widened Day 18's narrow, ambiguous 10-point sweep to a 40-point, log-spaced grid and reran it on mizer 3.1's second-order finite-volume scheme plus external diffusion. The post originally read that result backwards: it said the forward/backward gap closed. Re-reading the actual figure showed the opposite - the backward branch collapses to zero around `resource_decrease ≈ 0.02`, while the forward branch keeps a nonzero cycle out to `≈ 0.15–0.2`, a wide and unambiguous hysteresis loop, much bigger than Day 18's narrow `rd ≈ 0.008` ambiguity. The title, description, and body of that post have all been corrected to say so. The diffusion caveat stands either way: `ext_diffusion` was filled as a flat constant across all sizes rather than mizer's own allometric `w^(n+1)` shape, so the result (now: a wide hysteresis loop, not a closed gap) still needs to be reproduced with diffusion set up properly before it can be trusted.

Today picks that fix up directly, and then goes further: instead of only ever looking at each `resource_decrease` value's settled max/min biomass, it looks at biomass over time itself, to see the shape of the oscillation (or its absence) directly, and to watch the hysteresis play out step by step rather than reading it off a summary curve.

```{r}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse)
library(glue)
library(ggplot2)
library(colorRamps)
library(future.apply)
```

------------------------------------------------------------------------

## Setup

`make_second_order_params()` carries over the second-order finite-volume scheme from Day 20, but fixes the diffusion setup that post's "A Catch" section flagged as wrong. Instead of filling an `ext_diffusion` array with a flat constant, it now sets the baseline `D_ext` rate directly in `species_params` via `given_species_params()`.:

```{r}
#| message: false
#| warning: false
#| output: false
make_second_order_params <- function(lambda = 2.05, resource_decrease = 0.001,
                                     second_order = TRUE, ext_diff = 0.01) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  params <- newSingleSpeciesParams(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )
  r      <- getResourceRate(params) * resource_decrease
  params <- setResource(params, resource_rate = r,
                        resource_dynamics = "resource_semichemostat",
                        balance = TRUE)

  # given_species_params<-() is mizer's "stronger" setter: unlike
  # species_params<-(), it also triggers a recalculation of the derived
  # parameters, so setting the baseline D_ext rate here and letting mizer
  # recompute is the fix Day 20 flagged as still needed -- editing the
  # baseline rate and letting setExtDiffusion()'s w^(n+1) shape come from
  # mizer itself, rather than constant-filling the array by hand.
  if (second_order) {
    second_order_w(params) <- c(flux = "centred", bin_average = TRUE)
  }
  given_species_params(params)$D_ext <- ext_diff
  params
}
```

`rd_time_series` picks five `resource_decrease` values spanning Day 20's hysteresis region: deep in the oscillating regime, through the transition, and out past where the backward branch had collapsed.

```{r}
#| message: false
#| warning: false
#| output: false
rd_time_series <- c(0.001, 0.01, 0.05, 0.15, 0.3)
```

------------------------------------------------------------------------

## Experiment: Biomass Over Time, Run Independently

The first pass runs each `resource_decrease` value on its own, 300 years, starting from mizer's default initial state every time rather than carrying anything over between runs:

```{r}
#| message: false
#| warning: false
#| output: false
biomass_over_time <- lapply(rd_time_series, function(rd) {
  p   <- make_second_order_params(resource_decrease = rd)
  sim <- project(p, t_max = 300, dt = 0.1, t_save = 0.5,
                progress_bar = FALSE, effort = 0, method = "tr_bdf2")
  bm  <- getBiomass(sim)[, "Anchovy"]
  data.frame(time = as.numeric(names(bm)), biomass = as.numeric(bm),
            resource_decrease = factor(rd))
})
biomass_over_time_df <- bind_rows(biomass_over_time)
```

```{r}
#| message: false
#| warning: false
ggplot(biomass_over_time_df, aes(x = time, y = biomass, color = resource_decrease)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time", y = "Biomass",
       title = "Biomass over time for a few resource_decrease values",
       subtitle = "Flat line = settled fixed point; oscillating = limit cycle",
       color = "resource_decrease") +
  theme_minimal()
```

### A Backwards Signal, Read at Face Value

Read left to right in `resource_decrease`, the direction the plot above should show is backwards from what standard consumer-resource theory would predict: `rd = 0.001`, the resource growth rate cut to 0.1% of its default value, sustains the strongest, most persistent oscillation of the five, while `rd = 0.15` and `rd = 0.3`, where far more of the default resource rate survives, collapse to (near-)zero - the same direction the forward chain below settles into, reported there with actual numbers (`~0.055` at `rd = 0.05`, near-zero at `rd = 0.15`/`0.3`). More resource driving the population extinct, and a heavily depleted resource sustaining it, is the opposite of what cutting a resource base should do. That doesn't read like a real ecological result on its face; it reads like a sign or scaling error somewhere in how `resource_decrease`, the second-order scheme, and the diffusion fix above interact with the resource dynamics.

Whether that turns out to be a genuine (if backwards-looking) piece of dynamics or a parameterisation bug changes how it should be fixed, but not what needs to happen first: either way it's the same "characterise the bifurcation properly instead of eyeballing sweep plots" item already on the list under What's Next, so that's where this gets tracked down rather than treated as settled here.

This run also only ever approaches each value one way, from the same default starting point, so on its own it can't say whether the high-`rd` collapse is a genuine fixed point or just an artifact of that shared starting condition. It's the "normal" run the next section, a proper chained forward/backward pair, compares against.

------------------------------------------------------------------------

## Experiment: The Same Values, Chained Forward and Backward

The more useful version carries state between steps the same way Day 20's `fwd_so`/`bwd_so` sweep did: each `resource_decrease` value's run starts from the *previous* value's settled abundances, instead of resetting every time. `run_biomass_time_series()` does that, keeping the full biomass trace (not just the settled max/min) for every step, with time offset cumulatively across steps so each direction reads as one continuous trajectory:

```{r}
#| message: false
#| warning: false
#| output: false
run_biomass_time_series <- function(rd_seq, init_n = NULL, init_n_pp = NULL, t_run = 300) {
  state_n   <- init_n
  state_npp <- init_n_pp
  t_offset  <- 0
  segments  <- vector("list", length(rd_seq))

  for (i in seq_along(rd_seq)) {
    p <- make_second_order_params(resource_decrease = rd_seq[i])
    if (!is.null(state_n)) {
      p@initial_n[]    <- state_n
      p@initial_n_pp[] <- state_npp
    }
    sim <- project(p, t_max = t_run, dt = 0.1, t_save = 0.5,
                   progress_bar = FALSE, effort = 0, method = "tr_bdf2")
    bm   <- getBiomass(sim)[, "Anchovy"]
    tv   <- as.numeric(names(bm))
    last <- dim(sim@n)[1]

    state_n   <- sim@n[last, , ]
    state_npp <- sim@n_pp[last, ]

    segments[[i]] <- data.frame(time = tv + t_offset, biomass = as.numeric(bm),
                                resource_decrease = rd_seq[i])
    t_offset <- t_offset + max(tv)
  }
  list(df = bind_rows(segments), n_final = state_n, npp_final = state_npp)
}

fwd_ts <- run_biomass_time_series(rd_time_series)
bwd_ts <- run_biomass_time_series(rev(rd_time_series),
                                  init_n    = fwd_ts$n_final,
                                  init_n_pp = fwd_ts$npp_final)

biomass_fwd_bwd_df <- bind_rows(
  fwd_ts$df %>% mutate(direction = "Forward"),
  bwd_ts$df %>% mutate(direction = "Backward")
)
```

```{r}
#| message: false
#| warning: false
ggplot(biomass_fwd_bwd_df, aes(x = time, y = biomass, color = direction)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (cumulative across steps)", y = "Biomass",
       title = "Biomass over time: forward vs backward through the same resource_decrease values",
       subtitle = "Each segment holds resource_decrease fixed for 300 years before stepping to the next value",
       color = "Direction") +
  theme_minimal()
```

### What This Shows

Reading the five 300-year segments left to right:

**Forward** (`0.001 → 0.01 → 0.05 → 0.15 → 0.3`): strong limit-cycle oscillation at `rd = 0.001` (biomass swinging roughly `0.02–0.22`), settling almost immediately to a steady `~0.085` at `rd = 0.01`, then a lower steady `~0.055` at `rd = 0.05`, then collapsing to near-zero for both remaining segments (`rd = 0.15`, `0.3`).

**Backward** (`0.3 → 0.15 → 0.05 → 0.01 → 0.001`, starting from wherever forward ended up, already collapsed): stays collapsed near zero through *all three* of `rd = 0.3`, `0.15`, and `0.05`, only recovering to the same `~0.085` steady state once `rd` drops to `0.01`, and only re-entering the large oscillation once `rd` reaches `0.001`.

The disagreement at `rd = 0.05` is the headline: forward finds a healthy `~0.055` steady population there, backward is still fully collapsed. Which state the model is in at that `resource_decrease` value depends on which direction it was approached from, not on the value itself. That's the same wide hysteresis loop Day 20 found in the bifurcation diagram, now visible directly in the time series rather than inferred from a settled max/min summary.

------------------------------------------------------------------------

## Parked, Unfinished Threads

A few further ideas got started in `21_experiments.R` today but aren't in a state to draw conclusions from yet, and are left here as flagged rather than fixed.

### Reordering the Forward/Backward Chain

A draft ran the backward pass first (fresh, no carried-over state) and chained forward from backward's ending, the reverse of the convention used above and in Day 20:

```{r}
#| message: false
#| warning: false
#| output: false
run_rd_sweep_second_order <- function(rd_seq, init_n = NULL, init_n_pp = NULL,
                                      t_run = 600, lambda = 2.05,
                                      second_order = TRUE, label = "") {
  out       <- data.frame(rd = rd_seq, max_bm = NA_real_, min_bm = NA_real_,
                          bm_mean = NA_real_)
  state_n   <- init_n
  state_npp <- init_n_pp

  for (i in seq_along(rd_seq)) {
    p <- make_second_order_params(lambda = lambda, resource_decrease = rd_seq[i],
                                  second_order = second_order)
    if (!is.null(state_n)) {
      p@initial_n[]    <- state_n
      p@initial_n_pp[] <- state_npp
    }
    sim <- project(p, t_max = t_run, dt = 0.1, t_save = 0.5,
                   progress_bar = FALSE, effort = 0, method = "tr_bdf2")
    bm   <- getBiomass(sim)[, "Anchovy"]
    tv   <- as.numeric(names(bm))
    late <- bm[tv > t_run * 0.6]  # last 40% of the run -- the "settled down" window
    last <- dim(sim@n)[1]

    state_n   <- sim@n[last, , ]
    state_npp <- sim@n_pp[last, ]

    out$max_bm[i]  <- max(late)
    out$min_bm[i]  <- min(late)
    out$bm_mean[i] <- mean(late)
  }
  list(df = out, n_final = state_n, npp_final = state_npp)
}

# Same log-spaced grid as Day 20's wide sweep on the x-axis, but not a
# like-for-like rerun of it: this uses the fixed, allometric D_ext diffusion
# setup from this post's Setup section (Day 20's sweep used the flat-constant
# ext_diffusion fill) and a longer t_run = 600 rather than 300.
rd_seq_so <- exp(seq(log(0.0001), log(0.5), length.out = 40))

bwd_so <- run_rd_sweep_second_order(rev(rd_seq_so), label = "BWD (2nd order)")
fwd_so <- run_rd_sweep_second_order(rd_seq_so,
                                    init_n    = bwd_so$n_final,
                                    init_n_pp = bwd_so$npp_final,
                                    label     = "FWD (2nd order)")

fwd_so_df <- fwd_so$df
bwd_so_df <- bwd_so$df[order(bwd_so$df$rd), ]

bifurcation_df_so <- bind_rows(
  data.frame(rd = fwd_so_df$rd, biomass = fwd_so_df$max_bm, direction = "Forward", branch = "max"),
  data.frame(rd = fwd_so_df$rd, biomass = fwd_so_df$min_bm, direction = "Forward", branch = "min"),
  data.frame(rd = bwd_so_df$rd, biomass = bwd_so_df$max_bm, direction = "Backward", branch = "max"),
  data.frame(rd = bwd_so_df$rd, biomass = bwd_so_df$min_bm, direction = "Backward", branch = "min")
)
```

```{r}
#| message: false
#| warning: false
ggplot(bifurcation_df_so, aes(x = rd, y = biomass, color = direction, linetype = branch)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_x_log10() +
  labs(x = "resource_decrease", y = "Biomass",
       title = "Bifurcation diagram: backward-first chain (reordered convention)",
       subtitle = "Backward run fresh first, forward chained from its ending -- reverse of Day 20/21's usual order") +
  theme_minimal()
```

Whether this ordering matters for the shape of the loop hasn't been checked against the standard forward-first convention used above and in Day 20, so take this figure as a preliminary snapshot of what the swapped order produces, not yet a verified result.

## What's Next

- **Finish the Beverton-Holt attempt properly** so as to see whether stock-recruitment density dependence changes the size of the hysteresis loop found on Day 20/21.

- **Vary the diffusion rate** to see how the plot changes.

- **Keep working through *Nonlinear Dynamics and Chaos:*** This is to build the background needed to properly characterise the bifurcation (and the parked chaos-looking cell from Day 19), rather than continuing to read these sweep plots at face value.

- **Investigate the juvenile-biomass conundrum from Day 19:** Will do this via changing the juvenile slope ad increasing the size range of the species.
