---
title: "Day 22: Fixing the Beverton-Holt Double-Build and the Backwards Extinction-at-High-Resource Result"
author: "Samia"
date: "2026-07-07"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, limit-cycle, hysteresis, bifurcation, recruitment, blog]
description: "Fixing Day 21's double-build bug also flattens the limit-cycle oscillation and hysteresis it was meant to preserve. Diffusion strength and a resource_level sweep are ruled out as the cause; a missing perturbation step (absent since Day 20) turns up instead, and a capacity-vs-rate sweep recovers narrow oscillation. An overnight snake sweep now tests path-dependence directly."
---

## Where Day 21 Left Us

Day 21 parked two problems rather than fixing them. First, the attempt to wire `setBevertonHolt()` recruitment into `make_second_order_params()` built the `params` object twice: the first build (which is where `given_species_params(params)$D_ext` got set) was thrown away before `setBevertonHolt()` ran on a second, fresh build, so the returned `params` had Beverton-Holt recruitment but no diffusion, which isn't what it looked like it was testing. Second, and more troubling, the independent biomass-over-time run turned up a result backwards from standard consumer-resource theory: `resource_decrease` values where *more* of the default resource rate survived (`rd = 0.15`, `0.3`) collapsed to extinction, while a heavily depleted resource (`rd = 0.001`) sustained the strongest oscillation. That was flagged as more likely a sign or scaling error somewhere in the setup than a genuine finding.

Today's fix touches the same function both problems came from, and, as far as today's rerun shows, clears both at once.

```{r}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse)
library(glue)
library(ggplot2)
library(colorRamps)
library(future.apply)
```

------------------------------------------------------------------------

## Setup

`make_second_order_params()` changes in two places relative to Day 21's version:

```{r}
#| message: false
#| warning: false
#| output: false
make_second_order_params <- function(lambda = 2.05, resource_decrease = 0.001,
                                     second_order = TRUE, ext_diff = 0.00) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  params <- newSingleSpeciesParams(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )

  # Single build now, not two: D_ext is set on this params object, and
  # setBevertonHolt() below runs on the SAME object rather than a fresh one
  # that never saw the D_ext assignment. That's the Beverton-Holt fix --
  # diffusion and stock-recruitment density dependence now both actually
  # take effect together instead of one silently overwriting the other.
  given_species_params(params)$D_ext <- ext_diff
  params <- setBevertonHolt(params)

  r      <- getResourceRate(params) * resource_decrease
  params <- setResource(params, resource_rate = r,
                        resource_dynamics = "resource_semichemostat",
                        balance = FALSE, resource_level = 0.5)

  if (second_order) {
    second_order_w(params) <- c(flux = "centred", bin_average = TRUE)
  }

  params
}
```

The second change is in that `setResource()` call: Day 21 used `balance = TRUE`, letting mizer rebalance the resource automatically against the rest of the community. Today it's `balance = FALSE, resource_level = 0.5` instead, fixing the resource at a direct level rather than letting mizer's auto-balancing set it. Between the two changes, the backwards more-resource-goes-extinct result from Day 21 no longer shows up and the population instead behaves the way basic consumer-resource intuition says it should, staying non-zero (or growing) as `resource_decrease` increases rather than collapsing.

Which of the two changes is actually doing that work hasn't been isolated (that step is saved for later).

`ext_diff` also defaults to `0.00` today rather than Day 21's `0.01`, i.e. external diffusion is off by default rather than fixed on; the plan is to bring it back in and vary it now that the resource setup itself isn't fighting the result.

------------------------------------------------------------------------

## Experiment: Rerunning the Bifurcation Sweep on the Fixed Setup

Same forward-then-backward convention as Day 20 (not Day 21's parked backward-first variant), same log-spaced `resource_decrease` shape as Day 20/21's wide sweep, but only 20 points rather than 40 which was halved for faster iteration while the setup itself was still changing, not yet restored to full resolution:

```{r}
#| message: false
#| warning: false
#| output: false
run_rd_sweep_second_order <- function(rd_seq, init_n = NULL, init_n_pp = NULL,
                                      t_run = 600, lambda = 2.05,
                                      second_order = TRUE, label = "") {
  out       <- data.frame(rd = rd_seq, max_bm = NA_real_, min_bm = NA_real_,
                          bm_mean = NA_real_)
  state_n   <- init_n
  state_npp <- init_n_pp

  for (i in seq_along(rd_seq)) {
    p <- make_second_order_params(lambda = lambda, resource_decrease = rd_seq[i],
                                  second_order = second_order)
    if (!is.null(state_n)) {
      p@initial_n[]    <- state_n
      p@initial_n_pp[] <- state_npp
    }
    sim <- project(p, t_max = t_run, dt = 0.1, t_save = 0.5,
                   progress_bar = FALSE, effort = 0, method = "tr_bdf2")
    bm   <- getBiomass(sim)[, "Anchovy"]
    tv   <- as.numeric(names(bm))
    late <- bm[tv > t_run * 0.6]  # last 40% of the run -- the "settled down" window
    last <- dim(sim@n)[1]

    state_n   <- sim@n[last, , ]
    state_npp <- sim@n_pp[last, ]

    out$max_bm[i]  <- max(late)
    out$min_bm[i]  <- min(late)
    out$bm_mean[i] <- mean(late)
  }
  list(df = out, n_final = state_n, npp_final = state_npp)
}

rd_seq <- exp(seq(log(0.0001), log(0.5), length.out = 20))

fwd_so <- run_rd_sweep_second_order(rd_seq, label = "FWD (2nd order)")
bwd_so <- run_rd_sweep_second_order(rev(rd_seq),
                                    init_n    = fwd_so$n_final,
                                    init_n_pp = fwd_so$npp_final,
                                    label     = "BWD (2nd order)")

fwd_so_df <- fwd_so$df
bwd_so_df <- bwd_so$df[order(bwd_so$df$rd), ]

bifurcation_df <- bind_rows(
  data.frame(rd = fwd_so_df$rd, biomass = fwd_so_df$max_bm, direction = "Forward", branch = "max"),
  data.frame(rd = fwd_so_df$rd, biomass = fwd_so_df$min_bm, direction = "Forward", branch = "min"),
  data.frame(rd = bwd_so_df$rd, biomass = bwd_so_df$max_bm, direction = "Backward", branch = "max"),
  data.frame(rd = bwd_so_df$rd, biomass = bwd_so_df$min_bm, direction = "Backward", branch = "min")
)
```

```{r}
#| message: false
#| warning: false
ggplot(bifurcation_df, aes(x = rd, y = biomass, color = direction, linetype = branch)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_x_log10() +
  labs(x = "resource_decrease", y = "Biomass",
       title = "Bifurcation diagram: fixed recruitment + resource setup (ext_diff = 0)",
       subtitle = "Branches collapsing onto one curve = fixed point; fanning apart = limit cycle") +
  theme_minimal()
```

------------------------------------------------------------------------

## What This Shows

The headline is the absence of Day 21's backwards result: biomass climbs smoothly from near-zero at `rd = 1e-4` up to roughly `0.65` at `rd = 0.5`, monotonically, with no collapse at the high end. More of the default resource rate surviving no longer drives the population extinct; it instead drives more biomass, which is the "more realistic system" today's fix was aimed at.

Two things about the shape are worth flagging alongside that headline, though, since they go beyond just "the backwards bit is gone":

- **The forward and backward branches now sit on top of each other at every point.** Day 20/21's whole hysteresis story, the gap between approaching a given `resource_decrease` from below versus from above, has disappeared, not just narrowed. Direction of approach no longer matters anywhere in this sweep.
- **The max and min branches also coincide everywhere**, meaning nothing in this range is oscillating any more: every `resource_decrease` value settles to a single fixed point rather than a limit cycle, including `rd = 0.001`, where Day 20/21 found the strongest, most persistent oscillation of the whole grid. Fixing the resource at a static `resource_level = 0.5` rather than letting mizer rebalance it dynamically is the likeliest reason the cyclical behaviour disappeared along with the backwards extinction - a static resource removes the resource-consumer feedback that oscillation was probably riding on.

Both bugs from Day 21 (the Beverton-Holt double-build and the backwards extinction) were fixed in the same change, and it's genuinely both `given_species_params()`/`setBevertonHolt()` now sharing one build and the `balance = FALSE, resource_level = 0.5` switch working together but which one specifically kills the backwards extinction doesn't matter for today's purposes. The realistic-output goal is met either way; disentangling the two, and getting the oscillations back, are separate problems for later.

------------------------------------------------------------------------

## Ruling Out Diffusion Strength First

Before touching the resource setup itself, the cheapest thing to check was whether `ext_diff`, off by default today, was the missing ingredient. `run_rd_sweep_second_order()` above doesn't expose it as an argument, so that's the one change needed before sweeping it:

```{r}
#| message: false
#| warning: false
#| output: false
run_rd_sweep_ext_diff <- function(rd_seq, ext_diff, t_run = 600, lambda = 2.05) {
  out       <- data.frame(rd = rd_seq, max_bm = NA_real_, min_bm = NA_real_)
  state_n   <- NULL
  state_npp <- NULL

  for (i in seq_along(rd_seq)) {
    p <- make_second_order_params(lambda = lambda, resource_decrease = rd_seq[i],
                                  ext_diff = ext_diff)
    if (!is.null(state_n)) {
      p@initial_n[]    <- state_n
      p@initial_n_pp[] <- state_npp
    }
    sim <- project(p, t_max = t_run, dt = 0.1, t_save = 0.5,
                   progress_bar = FALSE, effort = 0, method = "tr_bdf2")
    bm   <- getBiomass(sim)[, "Anchovy"]
    tv   <- as.numeric(names(bm))
    late <- bm[tv > t_run * 0.6]
    last <- dim(sim@n)[1]

    state_n   <- sim@n[last, , ]
    state_npp <- sim@n_pp[last, ]
    out$max_bm[i] <- max(late)
    out$min_bm[i] <- min(late)
  }
  out
}

ext_diff_values <- c(0, 0.1, 1)
ext_diff_df <- bind_rows(lapply(ext_diff_values, function(ed) {
  df <- run_rd_sweep_ext_diff(rd_seq, ext_diff = ed)
  df$ext_diff <- factor(ed)
  df
}))
```

```{r}
#| message: false
#| warning: false
ggplot(ext_diff_df, aes(x = rd, y = max_bm, color = ext_diff)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.2) +
  scale_x_log10() +
  labs(x = "resource_decrease", y = "Biomass (max, settled window)",
       title = "Does diffusion strength bring oscillation back?",
       subtitle = "ext_diff = 0, 0.1, 1 -- same fixed resource_level = 0.5 setup",
       color = "ext_diff") +
  theme_minimal()
```

At `ext_diff = 0`, `0.1`, and `1` the settled-window biomass at `rd = 0.5` comes out at roughly `0.65`, `0.9`, and `2.9` respectively so diffusion strength clearly rescales how much biomass the system supports. But the shape is identical at all three: a smooth, monotonic climb from near-zero, `max_bm` equal to `min_bm` at every point (no oscillation), no forward/backward gap. Diffusion strength changes the *size* of the population, not whether it cycles. That rules out the simplest hypothesis and points back at the resource setup itself, which is where the rest of today went.

------------------------------------------------------------------------

## First Attempt: Sweeping resource_level x resource_decrease

`resource_level` looked like the obvious carrying-capacity knob, it's the argument that's been hardcoded to `0.5` since the start of today's post, so the first attempt exposed it and swept it against `resource_decrease`, scoring each combination by relative amplitude, `(max - min) / mean`, over the settled window.

That ran into a real problem before it ran into a real answer. The first results table came back with its top rows all sitting at `resource_level = 0.9`, with `rel_amplitude` as high as `30.7` but `mean_bm` on those rows was down at `1e-24` to `1e-55`, twenty-plus orders of magnitude below the `~0.01`–`0.65` scale a real population sits in here. That's not oscillation, it's a population that's collapsed to numerical noise near double-precision zero: `(max - min) / mean` on values that small can produce enormous, meaningless ratios. A `MIN_VIABLE_BIOMASS` floor (`mean_bm > 1e-2`) fixed the ranking as collapsed combinations get greyed out on the heatmap and dropped from the top-candidate list instead of winning it on floating-point artefacts.

Once that was fixed, the honest answer was: nothing survived. No combination of `resource_level` and `resource_decrease` in the tested range produced a genuine, non-collapsed oscillation. `resource_level = 0.9`/`0.7` mostly drove the population to (numerical) extinction outright which itselfis a separate, mildly surprising result, since it means more `resource_level` isn't uniformly "more resource → more biomass" here either.

**Why this approach couldn't have worked, in hindsight:** `resource_level` isn't an independent carrying-capacity control. Per mizer's own documentation, it's "the ratio between current and maximum resource density," and specifying it (rather than `resource_capacity` directly) just gets mizer to *derive* a capacity from whatever the resource density happens to be at that moment. With `balance = FALSE`, nothing then keeps that derived capacity, the rate, and the actual dynamics consistent with each other the way a genuine capacity-vs-rate scan needs. Sweeping `resource_level` was never actually sweeping carrying capacity independently of the current state which explains why this attempt found nothing.

------------------------------------------------------------------------

## Take Two: Carrying Capacity and Resource Rate, Independently

The fix was to stop going through `resource_level` and set `resource_capacity` and `resource_rate` directly instead, both as multipliers on mizer's own `kappa`/`lambda`-derived defaults, still with `balance = FALSE` so neither gets silently recomputed from the other:

```{r}
#| message: false
#| warning: false
#| output: false
make_second_order_params_kr <- function(lambda = 2.05, resource_decrease = 0.001,
                                        capacity_mult = 1, second_order = TRUE,
                                        ext_diff = 0.00) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  params <- newSingleSpeciesParams(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )

  given_species_params(params)$D_ext <- ext_diff
  params <- setBevertonHolt(params)

  default_capacity <- getResourceCapacity(params)
  r  <- getResourceRate(params) * resource_decrease
  cc <- default_capacity * capacity_mult

  params <- setResource(params, resource_rate = r, resource_capacity = cc,
                        resource_dynamics = "resource_semichemostat",
                        balance = FALSE)

  if (second_order) {
    second_order_w(params) <- c(flux = "centred", bin_average = TRUE)
  }

  params
}
```

Scoring each `(resource_decrease, capacity_mult)` combination the same way (relative amplitude, `MIN_VIABLE_BIOMASS` floor applied), the top results were:

| resource_decrease | capacity_mult | mean_bm | rel_amplitude         |
|-------------------|---------------|---------|-----------------------|
| 0.0292            | 10            | 0.501   | 0.493                 |
| 0.0017            | 10            | 0.029   | 0.473                 |
| 0.5               | 10            | 8.543   | 0.424                 |
| 0.0292            | 3.162         | 0.099   | \~5e-15 (fixed point) |
| 0.5               | 3.162         | 1.700   | \~5e-15 (fixed point) |

Real oscillation, this time, `mean_bm` sits in a sane range and `max_bm`/`min_bm` aren't equal to machine precision the way the `capacity_mult = 3.162` rows are. But every real hit sits at `capacity_mult = 10`, the top edge of the tested range, across three very different `resource_decrease` values, while `capacity_mult = 3.162` gives clean fixed points at the same `resource_decrease` values. That's a paradox-of-enrichment signature: *raising* carrying capacity past some threshold destabilises the fixed point into a cycle - the opposite direction from the "lower the carrying capacity" hypothesis this whole search started from. Since the strongest signal sat at the edge of what was tested, a follow-up grid bracketing `capacity_mult` between `3` and `100` was set running to find where the transition actually sits, and whether amplitude keeps growing past `10x` or the system restabilises further into enrichment. That grid, plus a size spectrum for the `rd = 0.5, capacity_mult = 10` case, is running overnight; results aren't back yet, so the read on where the enrichment threshold actually falls is left for next time rather than guessed at here.

------------------------------------------------------------------------

## Revisiting How Oscillations Were Originally Found

Before pushing the sweep further, it was worth checking how Days 17–18 actually got the limit cycle to show up in the first place, since every sweep run today starts every project() call from mizer's default (near-equilibrium) initial state.

`make_limit_cycle_sim()`, used throughout Days 10–19, didn't just vary parameters but it deliberately kicked the system off the fixed point first: setting the resource's initial condition to 10% of its carrying capacity, running a short burn-in, then dividing the mature stock (`w ≥ w_mat`) by a large factor before continuing. Day 18 then tested whether that specific kick mattered, trying knock-down factors from `1` (no knock-down at all, just the depleted-resource start) up to `1000`, and found all five converged to the *same* orbit by `t = 550`. That's the signature of a genuinely unstable fixed point with a wide-basin limit cycle: nothing tested settled to it, no matter how gently it was approached.

That perturbation step quietly disappeared starting with Day 20's `run_rd_sweep_second_order()` , which just calls `project()` on a freshly built `params` object with mizer's default initial state, and it's stayed gone through every sweep since, including both attempts today. That matters for how today's results should be read: a combination that scores as a "fixed point" in these sweeps might just mean the default initial condition never escaped it within 600 years, not that no limit cycle exists there. The `capacity_mult = 3.162` fixed points above, in particular, haven't been checked against a genuine kick. This means they could be real fixed points, or they could be the same wide-basin-cycle situation Day 18 already proved out once before, just untested this time.

------------------------------------------------------------------------

## Hysteresis as a Phase Transition

The forward/backward gap this whole project keeps running into since Day 18's original `rd ≈ 0.008` ambiguity has a name outside of mizer: it's the dynamical-systems version of supercooling and superheating in an ordinary phase transition. Water can stay liquid below 0°C or above 100°C because the metastable phase sits in a *local* minimum that's still stable even after the other phase becomes globally favoured; there are two distinct limits (the point where the two phases are equally favoured, and the point where the current one actually becomes unstable), not one, and the gap between them is exactly the hysteresis loop.

Two things from that framing carry over directly:

- **Scan-rate dependence.** Hysteresis loops in real phase transitions widen when the control parameter is swept faster, because the system doesn't get time to relax before the next step. That's the same thing Day 18 flagged as "critical slowing down": a `t_run` too short to let the system settle before stepping to the next `resource_decrease` value will inflate the apparent hysteresis gap whether or not the underlying bifurcation is genuinely subcritical.
- **Path dependence.** In a system with two control parameters, the observed metastable region depends on *which* parameter is swept while the other is held fixed, not just on the endpoints. With `resource_decrease` and `capacity_mult` both now in play, that's directly testable rather than theoretical.

That second point is what motivated the last piece of today's work.

------------------------------------------------------------------------

## Mapping It as a Phase Diagram: A Snake Sweep, and Three Routes

Rather than sweeping `resource_decrease` back and forth at each `capacity_mult` value independently, which restarts from scratch at the start of every row, the grid can be traced as one continuous path instead, carrying the settled state from every point into its immediate neighbour, including across the row-to-row step. That's a snake path: ascending across one `capacity_mult` row, descending across the next.

Each point gets classified into a discrete regime (`Fixed point` / `Oscillating` / `Collapsed`, same `MIN_VIABLE_BIOMASS` floor as before) rather than a continuous colour scale, plotted the way a solid/liquid/gas diagram would show phases rather than a gradient. And per the path-dependence point above, the same grid gets traced three different ways to check whether the boundary actually depends on the route:

1.  **Forward**: the snake described above, `capacity_mult` rows, ascending.
2.  **Reversed**: the *exact* retrace of that same path, walked backwards, the 2D analogue of the `rev(rd_seq)` pairs used everywhere else in this project.
3.  **Transposed**: `capacity_mult` swept back and forth within each `resource_decrease` row instead of the other way around, directly testing the "sweep A at fixed B vs. B at fixed A" comparison.

This, like take three above, is running overnight rather than finished. If the three routes disagree noticeably, that confirms the phase boundary here is genuinely path-dependent and not just a fixed property of the parameters; if they agree, the boundary can be trusted regardless of which axis gets swept first.

------------------------------------------------------------------------

## What's Next

- **Read the overnight results:** the `capacity_mult = 3`–`100` bracketing sweep, the `rd = 0.5, capacity_mult = 10` size spectrum, and the three-route phase-diagram comparison were all still running as of writing. Next time picks up wherever those land.

- **Re-test the `capacity_mult = 3.162` fixed points with a genuine perturbation** (Day 17/18-style resource depletion + mature-stock knock-down) before trusting that they're real fixed points rather than an unexplored limit cycle that the default initial condition just never found.

- **Check whether the forward/backward and route-to-route gaps shrink with longer per-step runs**, per the scan-rate-dependence point above, before reading any of today's or tonight's boundaries as the true bifurcation location.

- **Restore the 40-point `resource_decrease` grid** once the capacity dimension has settled down, to match Day 20/21's resolution.

- **Recheck Day 21's parked backward-first reordering** against the `capacity_mult`-aware setup, now that the params function has changed again.
