---
title: "Day 20: A Wider Sweep Reveals a Much Bigger Hysteresis Gap - But the Diffusion Setup Needs Redoing First"
author: "Samia"
date: "2026-07-03"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, limit-cycle, bifurcation, hysteresis, numerical-methods, diffusion, blog]
description: "Widening Day 18's narrow, ambiguous resource_decrease sweep and rebuilding it on mizer 3.1's second-order scheme reveals a wide hysteresis loop, not a resolved bifurcation - though a flawed diffusion setup means the result needs to be reproduced before it can be trusted."
---

## Where Day 19 Left Us

Day 19 closed with four threads left open: plot maximum and minimum separately instead of collapsing to max − min, to get an actual bifurcation-diagram view of the amplitude data; try a different perturbation type; keep working through Strogatz's *Nonlinear Dynamics and Chaos*; and figure out why so much biomass piles up at juvenile sizes. Today picks up the first thread, and in the process pulls in two pieces of new information: mizerExperimental's `biomass_callback()` (for watching a sweep run live) and mizer 3.1's new second-order finite-volume scheme, `second_order_w()`.

```{r}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse)
library(glue)
library(ggplot2)
library(colorRamps)
library(future.apply)
```

------------------------------------------------------------------------

## Setup

The baseline `make_params()` is unchanged from Day 18/19. Alongside it, `make_second_order_params()` builds the same Anchovy params but with mizer 3.1's second-order finite-volume scheme switched on (`second_order_w(params) <- c(flux = "centred", bin_average = TRUE)`, cutting size-discretisation error from O(Δw) to O(Δw²)) and a background external diffusion added via `setExtDiffusion()`.

```{r}
#| message: false
#| warning: false
#| output: false
make_params <- function(lambda = 2.05, resource_decrease = 0.001,
                        beta = NULL, sigma = NULL, balance = TRUE) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  args <- list(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )
  if (!is.null(beta))  args$beta  <- beta
  if (!is.null(sigma)) args$sigma <- sigma

  params <- do.call(newSingleSpeciesParams, args)
  r      <- getResourceRate(params) * resource_decrease
  setResource(params, resource_rate = r,
             resource_dynamics = "resource_semichemostat", balance = balance)
}

make_second_order_params <- function(lambda = 2.05, resource_decrease = 0.001,
                                     second_order = TRUE, ext_diff = 0) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  params <- newSingleSpeciesParams(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )
  r      <- getResourceRate(params) * resource_decrease
  params <- setResource(params, resource_rate = r,
                        resource_dynamics = "resource_semichemostat",
                        balance = TRUE)

  if (second_order) {
    second_order_w(params) <- c(flux = "centred", bin_average = TRUE)
  }

  # setExtDiffusion() needs ext_diffusion as a species x size array, not a
  # bare scalar -- getDiffusion(params) is already validly shaped (all zero,
  # since no ext diffusion is set yet), so it's used as a dim/dimnames
  # template and filled with the scalar ext_diff. This turns out to be the
  # wrong way to do this -- see "A Catch" below.
  diffusion_template  <- getDiffusion(params)
  ext_diffusion_array <- array(ext_diff, dim = dim(diffusion_template),
                               dimnames = dimnames(diffusion_template))
  params <- setExtDiffusion(params, ext_diffusion = ext_diffusion_array)
  params
}

run_rd_sweep_second_order <- function(rd_seq, init_n = NULL, init_n_pp = NULL,
                                      t_run = 300, lambda = 2.05,
                                      second_order = TRUE, ext_diff = 0, label = "") {
  out       <- data.frame(rd = rd_seq, max_bm = NA_real_, min_bm = NA_real_,
                          bm_mean = NA_real_)
  state_n   <- init_n
  state_npp <- init_n_pp

  for (i in seq_along(rd_seq)) {
    p <- make_second_order_params(lambda = lambda, resource_decrease = rd_seq[i],
                                  second_order = second_order, ext_diff = ext_diff)
    if (!is.null(state_n)) {
      p@initial_n[]    <- state_n
      p@initial_n_pp[] <- state_npp
    }
    sim <- project(p, t_max = t_run, dt = 0.1, t_save = 0.5,
                   progress_bar = FALSE, effort = 0, method = "tr_bdf2")
    bm   <- getBiomass(sim)[, "Anchovy"]
    tv   <- as.numeric(names(bm))
    late <- bm[tv > t_run * 0.6]
    last <- dim(sim@n)[1]

    state_n   <- sim@n[last, , ]
    state_npp <- sim@n_pp[last, ]

    out$max_bm[i]  <- max(late)
    out$min_bm[i]  <- min(late)
    out$bm_mean[i] <- mean(late)
  }
  list(df = out, n_final = state_n, npp_final = state_npp)
}
```

------------------------------------------------------------------------

## Experiment: A Much Wider, Log-Spaced Hysteresis Sweep

Day 18's forward/backward sweep only covered `resource_decrease` = 0.001–0.01, ten points on a linear grid, and left an unresolved gap between the forward and backward transition points around `rd ≈ 0.008`. Today's sweep widens that considerably: 40 points, log-spaced rather than linear, spanning 0.0001–0.5, which is 3.7 orders of magnitude, at roughly the same per-decade resolution as Day 18's narrower grid rather than thinning out at the low end.

```{r}
#| message: false
#| warning: false
#| output: false
rd_seq <- exp(seq(log(0.0001), log(0.5), length.out = 40))

fwd_so <- run_rd_sweep_second_order(rd_seq, label = "FWD (2nd order)")
bwd_so <- run_rd_sweep_second_order(rev(rd_seq),
                                    init_n    = fwd_so$n_final,
                                    init_n_pp = fwd_so$npp_final,
                                    label     = "BWD (2nd order)")

fwd_so_df <- fwd_so$df
bwd_so_df <- bwd_so$df[order(bwd_so$df$rd), ]
```

Max and min are plotted separately rather than collapsed to `amp = max - min`, per Day 19's carried-over What's Next item, on a log x-axis since the grid now spans so many orders of magnitude:

```{r}
#| message: false
#| warning: false
bifurcation_df <- bind_rows(
  data.frame(rd = fwd_so_df$rd, biomass = fwd_so_df$max_bm, direction = "Forward", branch = "max"),
  data.frame(rd = fwd_so_df$rd, biomass = fwd_so_df$min_bm, direction = "Forward", branch = "min"),
  data.frame(rd = bwd_so_df$rd, biomass = bwd_so_df$max_bm, direction = "Backward", branch = "max"),
  data.frame(rd = bwd_so_df$rd, biomass = bwd_so_df$min_bm, direction = "Backward", branch = "min")
)

ggplot(bifurcation_df, aes(x = rd, y = biomass, color = direction, linetype = branch)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_x_log10() +
  labs(x = "resource_decrease", y = "Biomass",
       title = "Bifurcation diagram: second-order scheme + ext_diffusion, wide log-spaced sweep",
       subtitle = "Branches collapsing onto one curve = fixed point; fanning apart = limit cycle") +
  theme_minimal()
```

------------------------------------------------------------------------

## What the Wider Sweep Shows

Across the full 0.0001–0.5 range, the forward and backward max/min branches agree at both ends - collapsing onto the same fixed point below rd ≈ 0.005 and tracking similar cycle amplitude through the middle of the range - but they diverge sharply toward the top: the backward branch's cycle collapses to zero around rd ≈ 0.02, while the forward branch keeps a nonzero cycle going all the way out to rd ≈ 0.15–0.2 before it, too, drops to zero. That's a wide stretch, nearly a full order of magnitude in rd, where the two directions disagree about whether the system is still oscillating at all. Read at face value, this is a much clearer and much wider hysteresis loop than Day 18's narrow, ambiguous gap around rd ≈ 0.008, not a resolution of it.

That widening sharpens Day 18's open question rather than settling it, and there's one further problem with how today's sweep was actually built.

------------------------------------------------------------------------

## A Catch: the External Diffusion Was Set Up Wrong

`make_second_order_params()` builds its `ext_diffusion` array by taking `getDiffusion(params)`'s shape as a template and filling every single element with the same scalar, `ext_diff`. That's not what mizer's own external diffusion is supposed to look like. mizer's default formula is

$$D_{\text{ext},i}(w) = D_{\text{ext},i} \cdot w^{n_i + 1}$$

which is a per-species *baseline rate* (`D_ext` in `species_params`) that gets scaled allometrically across the size spectrum by `w` raised to the growth exponent plus one. Filling the whole array with one constant instead throws that shape away entirely: rather than a small diffusion term at small `w` growing smoothly into a much larger one at large `w` (or vice versa, depending on the exponent), every size class got exactly the same value regardless of body weight. Whatever `ext_diff = 0.001` produced today is not "external diffusion at a background level of 0.001" in any sense that connects back to mizer's own model of what diffusion means physically ; it's closer to an arbitrary, non-physical size-independent forcing term that happened to share a value with the intended baseline rate.

That means today's headline result, the wide hysteresis loop that opened up under the wider sweep, was generated under a diffusion setup that doesn't mean what it was intended to mean. It's a striking result, and the direction (second-order scheme, wider range, gap widens rather than narrows) is plausible on its own terms. But it needs to be reproduced with `ext_diffusion` set up properly, editing the baseline `D_ext` rate in `species_params` and letting `setExtDiffusion()` derive the `w^(n+1)` array itself, rather than constant-filling, before it can be trusted as showing what a *physically sensible* level of external diffusion actually does to the bifurcation.

------------------------------------------------------------------------

## Theoretical Grounding, Continued

Work through Strogatz's *Nonlinear Dynamics and Chaos* is ongoing, still centred on sections 8.3–8.7 (global bifurcations, hysteresis, quasiperiodicity, Poincaré maps) which is the exact toolkit needed to properly characterise what today's sweep is showing, rather than reading a collapsed-onto-one-curve plot at face value.

------------------------------------------------------------------------

## What's Next

- **Redo the external diffusion setup properly:** Instead of filling `ext_diffusion` with a flat constant across every size class, edit the baseline `D_ext` rate in `species_params` and let `setExtDiffusion()` derive mizer's intended $D_{\text{ext},i}(w) = D_{\text{ext},i} \cdot w^{n_i+1}$ shape itself. Then rerun the wide, log-spaced sweep to see whether the wide-hysteresis result survives under a physically sensible diffusion term, rather than the constant-array version used today.

- **Keep working through *Nonlinear Dynamics and Chaos:*** This is to build the background needed to properly characterise the bifurcation (and the parked chaos-looking cell from Day 19), rather than continuing to read these sweep plots at face value.

- **Change the juvenile slope, starting by reducing it:** Investigating the juvenile-biomass conundrum from Day 19: does flattening the size-spectrum slope specifically over the juvenile size range change the juvenile:adult standing-stock ratio, or is the imbalance insensitive to it?

- **Increase the size range of the species:** Investigating the same juvenile phenomenon from a different angle: does giving individuals more room to grow through (a wider `w_min`–`w_max` span) change how much biomass piles up at small sizes, or does the same lopsided ratio persist regardless of how much size range is available?
