---
title: "Day 26: Closing Day 25's Loose Ends, and a Delay That Actually Causes a Hopf Bifurcation"
author: "Samia"
date: "2026-07-13"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, juvenile-pileup, alpha, time-delay, hopf-bifurcation, predator-prey, blog]
description: "Three Day 25 loose ends closed: the juvenile-pileup alpha transition is pinned to (0.105, 0.178), the gamma = 305.8 dip is traced to an unsettled oscillation rather than a real effect, and the tau = 40/80 amplitude climb is confirmed to decay to zero. Separately, delaying the prey's self-limitation (rather than the predator's conversion) is shown to produce a genuine Hopf bifurcation, though its amplitude drives prey biomass unphysically negative past onset; a third variant, delaying only the predator's intake term, derives a real threshold on D but doesn't show the predicted oscillation within the delays tested."
---

## Where Day 25 Left Us

Day 25 closed the juvenile-pileup sweep and ruled out a Hopf bifurcation for the time-delay predator-prey model, but left three threads open: pin down $\alpha$'s transition window, check whether the $\gamma=305.8$ dip was real or a transient, and confirm the $\tau=40/80$ amplitude climb actually decays to zero. All three are closed today, in `26_experiments.R`. A further question ( if delay usually destabilises systems, why couldn't it here? ) led to re-deriving the Hopf condition for two other places the delay could sit, one of which does produce one.

------------------------------------------------------------------------

## Setup

```{r libraries}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse)
library(glue)
library(ggplot2)
library(colorRamps)
library(future)
library(scales)
library(deSolve)
```

Same anchovy builder and the same `catchable_fraction_at()`/`run_param_sweep()` pair from Day 25, carried over unchanged so today's refinements are directly comparable to the sweeps that found the transition and the dip in the first place:

```{r builder}
#| message: false
#| warning: false
#| output: false
make_second_order_params_kr <- function(lambda = 2.05, resource_decrease = 0.001,
                                        capacity_mult = 1, second_order = TRUE,
                                        ext_diff = 0.00, alpha = 0.1, gamma = 750,
                                        ks = 0, kappa_override = NULL) {
  a0    <- 100
  kappa <- if (is.null(kappa_override)) a0 * exp(-6.9 * (lambda - 1)) else kappa_override
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  params <- newSingleSpeciesParams(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = alpha, gamma = gamma, ks = ks
  )

  given_species_params(params)$D_ext <- ext_diff
  params <- setBevertonHolt(params)

  default_capacity <- getResourceCapacity(params)
  r  <- getResourceRate(params) * resource_decrease
  cc <- default_capacity * capacity_mult

  params <- setResource(params, resource_rate = r, resource_capacity = cc,
                        resource_dynamics = "resource_semichemostat",
                        balance = FALSE)

  if (second_order) {
    second_order_w(params) <- c(flux = "centred", bin_average = TRUE)
  }

  params
}

catchable_fraction_at <- function(params, t_max = 600) {
  sim <- project(params, t_max = t_max, dt = 0.1, t_save = 0.5,
                 progress_bar = FALSE, effort = 0, method = "tr_bdf2")
  last  <- dim(sim@n)[1]
  w_mat <- params@species_params$w_mat[1]
  w     <- sim@params@w
  dw    <- sim@params@dw

  bm_density   <- sim@n[last, 1, ] * w * dw
  total_bm     <- sum(bm_density)
  catchable_bm <- sum(bm_density[w >= w_mat])

  data.frame(catchable_fraction = catchable_bm / total_bm, error = NA_character_)
}

run_param_sweep <- function(param_name, param_seq, capacity_mult = 10,
                            resource_decrease = 0.001, t_max = 600) {
  bind_rows(lapply(param_seq, function(v) {
    args <- list(capacity_mult = capacity_mult, resource_decrease = resource_decrease)
    args[[param_name]] <- v
    result <- tryCatch({
      p <- do.call(make_second_order_params_kr, args)
      catchable_fraction_at(p, t_max = t_max)
    }, error = function(e) {
      data.frame(catchable_fraction = NA_real_, error = conditionMessage(e))
    })
    if (!is.na(result$error)) {
      warning(sprintf("%s = %.4g: %s", param_name, v, result$error))
    }
    data.frame(param = param_name, value = v,
               catchable_fraction = result$catchable_fraction,
               error = result$error)
  }))
}
```

------------------------------------------------------------------------

## Pinning Down Alpha's Transition Window

Day 25's 15-point $\alpha$ sweep found the catchable fraction jumps roughly 8 orders of magnitude between $\alpha=0.057$ and $\alpha=0.28$, but the grid wasn't dense enough to say where the steep part starts and ends. Same move as Day 24's `capacity_mult` refinement: 25 log-spaced points zoomed into $[0.05, 0.3]$, at the same `t_run = 600` for direct comparability.

```{r alpha-refine}
#| message: false
#| warning: false
alpha_seq_refine <- exp(seq(log(0.05), log(0.3), length.out = 25))
alpha_refine_df  <- run_param_sweep("alpha", alpha_seq_refine)
```

Unlike `capacity_mult`'s collapse bifurcation, there's no natural alive/dead split here. `catchable_fraction` rises smoothly from near-zero to a plateau, not a bistable jump. So the window is bracketed with two threshold crossings: where the climb clearly starts (under 0.1%) and where it's clearly finished (over 10%):

```{r find-crossing}
#| message: false
#| warning: false
#| output: false
find_crossing <- function(df, threshold, value_col = "value",
                          metric_col = "catchable_fraction") {
  df    <- df %>% arrange(.data[[value_col]])
  below <- df %>% filter(.data[[metric_col]] < threshold)  %>% pull(.data[[value_col]])
  above <- df %>% filter(.data[[metric_col]] >= threshold) %>% pull(.data[[value_col]])
  data.frame(
    threshold   = threshold,
    below_last  = if (length(below) > 0) max(below) else NA_real_,
    above_first = if (length(above) > 0) min(above) else NA_real_
  )
}

LOW_THRESHOLD  <- 1e-3
HIGH_THRESHOLD <- 0.10

alpha_crossings <- bind_rows(
  find_crossing(alpha_refine_df %>% filter(is.na(error)), LOW_THRESHOLD)  %>% mutate(edge = "climb starts"),
  find_crossing(alpha_refine_df %>% filter(is.na(error)), HIGH_THRESHOLD) %>% mutate(edge = "climb ends")
)
```

```{r alpha-plot}
#| message: false
#| warning: false
ggplot(alpha_refine_df %>% filter(is.na(error)), aes(x = value, y = catchable_fraction)) +
  geom_line(color = "#4C72B0", linewidth = 1) +
  geom_point(color = "#4C72B0", size = 1.5) +
  geom_hline(yintercept = c(LOW_THRESHOLD, HIGH_THRESHOLD), linetype = "dotted", color = "grey40") +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "grey40") +
  scale_y_log10(labels = scales::label_scientific()) +
  scale_x_log10() +
  labs(x = "alpha (assimilation efficiency)", y = "Catchable fraction (biomass at or above w_mat)",
       title = "Refined alpha sweep: pinning down the transition window",
       subtitle = "25 points over [0.05, 0.3] -- dotted = threshold crossings; dashed = anchovy's own alpha = 0.1") +
  theme_minimal()
```

The steep part of the climb is compressed into a narrow window, $\alpha \approx 0.105\text{–}0.178$ which is much tighter than Day 25's coarse "0.057 to 0.28." The anchovy's own $\alpha = 0.1$ sits at $0.0237\%$ catchable fraction, just below the 0.1% "climb starts" threshold. This confirms Day 25's "foot of the climb" read with an actual number.

------------------------------------------------------------------------

## Is the Gamma = 305.8 Dip Real, or an Under-Settled Transient?

Day 25's 15-point $\gamma$ sweep found one point breaking monotonicity. This code block reruns that point and its neighbours (located dynamically off the original grid) at increasingly long `t_run`.

```{r gamma-dip-setup}
#| message: false
#| warning: false
#| output: false
gamma_full_seq   <- exp(seq(log(100), log(5000), length.out = 15))
dip_idx          <- which.min(abs(gamma_full_seq - 305.8))
neighbour_idx    <- unique(pmax(1, pmin(length(gamma_full_seq), c(dip_idx - 1, dip_idx, dip_idx + 1))))
gamma_check_vals <- gamma_full_seq[neighbour_idx]

t_run_check_seq <- c(600, 1200, 2400, 4800)

gamma_dip_check_df <- bind_rows(lapply(t_run_check_seq, function(tr) {
  bind_rows(lapply(gamma_check_vals, function(g) {
    p   <- make_second_order_params_kr(gamma = g, capacity_mult = 10, resource_decrease = 0.001)
    res <- catchable_fraction_at(p, t_max = tr)
    data.frame(gamma = g, t_run = tr, catchable_fraction = res$catchable_fraction)
  }))
}))

dip_persistence <- gamma_dip_check_df %>%
  group_by(t_run) %>%
  arrange(gamma, .by_group = TRUE) %>%
  summarise(
    dip_still_present = catchable_fraction[2] < catchable_fraction[1] &
                        catchable_fraction[2] < catchable_fraction[3],
    .groups = "drop"
  )
```

```{r gamma-dip-plot}
#| message: false
#| warning: false
ggplot(gamma_dip_check_df, aes(x = t_run, y = catchable_fraction, color = factor(round(gamma, 1)))) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_x_log10() +
  scale_y_log10(labels = scales::label_scientific()) +
  labs(x = "t_run", y = "Catchable fraction",
       title = "Is the gamma = 305.8 dip real or an under-settled transient?",
       subtitle = "Same point and its two neighbours, rerun at increasing t_run", color = "gamma") +
  theme_minimal()
```

This isn't "the dip shrinks and disappears" (critical slowing down) or "the dip holds steady" (a real feature) - it flips non-monotonically, and at `t_run = 2400` the ordering of all three points scrambles, not just the middle one. That's not what a converging transient looks like.

The likely explanation: `catchable_fraction_at()` reads a single snapshot at the last saved timestep, not an average over a settled window. If the system at this operating point (`capacity_mult = 10`, `resource_decrease = 0.001`, the fixed baseline every juvenile-pileup sweep since Day 24 has used) is still gently oscillating rather than sitting at a true fixed point, varying `t_max` just samples different *phases*, not different amounts of settling time, which is consistent with Day 23's refined $[3,10]$ grid finding a reversal from fixed-point to oscillating right around `capacity_mult = 10`. That noise floor (values bouncing 2–3x between reruns) is negligible next to $\alpha$'s 27.5-order-of-magnitude range, but comparable to or larger than $\gamma$'s and $\kappa$'s entire reported ranges, so those "flat"/"modest effect" Day 25 conclusions may be sampling noise rather than a real parameter effect. Checking directly whether this operating point is actually oscillating is the natural follow-up.

------------------------------------------------------------------------

## Confirming the Amplitude Decay for the Conversion-Delay Model

Day 25's time-delay model (delay on the predator's conversion of intake into biomass) was proven analytically to never support a Hopf bifurcation, but `rel_amplitude_P` was still climbing with $\tau$ across the tested range (up to 0.25 at $\tau=40$), which the algebra says must eventually decay rather than plateau. Reproducing Day 25's model first to confirm the base case, then extending `t_max` well past the original 300:

```{r delay-reproduce}
#| message: false
#| warning: false
#| output: false
predator_prey_params_delay <- list(D = 1, a = 1, e = 0.5, m = 0.3, K = 1)

check_hopf_feasibility <- function(parms = predator_prey_params_delay) {
  D <- parms$D; a <- parms$a; e <- parms$e; m <- parms$m; K <- parms$K
  N_star <- m / (e * a)
  P_star <- D * (K - N_star) / (a * N_star)
  alpha  <- D + a * P_star
  beta   <- a * N_star
  gamma  <- e * a * P_star
  data.frame(N_star = N_star, P_star = P_star, alpha = alpha, beta = beta,
             gamma = gamma, beta_gamma = beta * gamma, two_m_alpha = 2 * m * alpha,
             hopf_possible = (beta * gamma) > (2 * m * alpha))
}
hopf_check <- check_hopf_feasibility()

predator_prey_rhs_delay <- function(t, state, parms) {
  N <- max(state[1], 0); P <- max(state[2], 0)
  D <- parms[["D"]]; a <- parms[["a"]]; e <- parms[["e"]]
  m <- parms[["m"]]; K <- parms[["K"]]; tau <- parms[["tau"]]
  N0 <- parms[["N0"]]; P0 <- parms[["P0"]]
  if (t <= tau) { lag_N <- N0; lag_P <- P0
  } else { lag <- lagvalue(t - tau); lag_N <- max(lag[1], 0); lag_P <- max(lag[2], 0) }
  dN <- D * (K - N) - a * N * P
  dP <- e * a * lag_N * lag_P - m * P
  list(c(dN, dP))
}

run_predator_prey_delay_long <- function(tau, t_max_long, state0 = c(N = 1, P = 0.1),
                                         parms = predator_prey_params_delay) {
  parms$tau <- tau; parms$N0 <- state0[["N"]]; parms$P0 <- state0[["P"]]
  as.data.frame(dede(y = state0, times = seq(0, t_max_long, by = 0.1),
                     func = predator_prey_rhs_delay, parms = parms))
}

amplitude_at_cutoff <- function(ts_df, t_max_effective) {
  late <- ts_df[ts_df$time > t_max_effective * 0.6 & ts_df$time <= t_max_effective, ]
  data.frame(t_max = t_max_effective, rel_amplitude_P = (max(late$P) - min(late$P)) / mean(late$P))
}

tau_check_seq   <- c(40, 80)
t_max_check_seq <- c(300, 600, 1200, 2400, 4800, 9600)

ts_long_by_tau <- setNames(
  lapply(tau_check_seq, function(tau) run_predator_prey_delay_long(tau, max(t_max_check_seq))),
  tau_check_seq
)

decay_check_df <- bind_rows(lapply(tau_check_seq, function(tau) {
  ts_long <- ts_long_by_tau[[as.character(tau)]]
  bind_rows(lapply(t_max_check_seq, function(tm) amplitude_at_cutoff(ts_long, tm) %>% mutate(tau = tau)))
}))
```

```{r decay-plot}
#| message: false
#| warning: false
ggplot(decay_check_df, aes(x = t_max, y = rel_amplitude_P, color = factor(tau))) +
  geom_line(linewidth = 1) + geom_point(size = 1.5) +
  scale_x_log10() + scale_y_log10(labels = scales::label_scientific()) +
  labs(x = "t_max", y = "Relative amplitude of P, late window",
       title = "Does rel_amplitude_P decay to ~0 given enough time, or plateau?",
       subtitle = "tau = 40 and 80, re-windowed at longer cutoffs -- analytic prediction: decays to 0",
       color = "tau") +
  theme_minimal()
```

Both $\tau$ values decay across 15 orders of magnitude to machine-precision zero, which is clearly not a plateau. An endpoint check (distance from the trajectory's final $N,P$ to the analytic equilibrium) confirms it directly: both land within $3.7\times10^{-15}$ of $(N^*,P^*)$ by `t_max=9600`. That closes Day 25's third loose end: the climbing amplitude was an unfinished monotonic relaxation, not a real or growing oscillation, exactly as predicted.

------------------------------------------------------------------------

## Why Couldn't Delay Cause an Instability Here?

Day 25's model delays the predator's *conversion* of intake into biomass, but leaves the prey's own self-limitation instantaneous:

$$\begin{aligned}
\frac{dN}{dt} &= D(K - N(t)) - aN(t)P(t) \\
\frac{dP}{dt} &= eaN(t-\tau)P(t-\tau) - mP(t)
\end{aligned}$$

**Step 1: Linearise.** Substituting $n,p\sim e^{\lambda t}$ around equilibrium gives the characteristic equation, with $\alpha=D+aP^*$, $\beta=aN^*$, $\gamma=eaP^*$:

$$(\lambda+\alpha)\left(\lambda+m-me^{-\lambda\tau}\right) + \beta\gamma e^{-\lambda\tau} = 0$$

**Step 2: Find the Hopf condition.** Setting $\lambda=i\omega$ and matching $|P(i\omega)|=|Q(i\omega)|$ reduces this to a quadratic in $u=\omega^2$:

$$u^2 + \alpha^2 u + \beta\gamma\left(2m\alpha - \beta\gamma\right) = 0$$

By Vieta's formulas, the sum of the roots is $-\alpha^2$ (always negative), so a positive root exists only if the *product*, $\beta\gamma(2m\alpha-\beta\gamma)$, is negative, i.e. $\beta\gamma>2m\alpha$.

**Step 3: Check whether that can ever hold.** Substituting and simplifying (using $eaN^*=m$):

$$\beta\gamma - 2m\alpha = -amP^* - 2mD$$

Factoring out $-mD$ (after substituting $P^*=D(K-N^*)/(aN^*)$) gives a clean closed form:

$$\beta\gamma - 2m\alpha = -\frac{mD(K+N^*)}{N^*}$$

Every quantity in that bracket is positive, so this can never be positive for *any* positive $D,a,e,m,K$. The undelayed prey renewal term $D$ is a restoring force acting at full strength regardless of $\tau$, strong enough here to prevent any Hopf bifurcation.

That raised an obvious question: what if the delay sat somewhere else?

------------------------------------------------------------------------

## A Model That Does Support a Hopf Bifurcation: Delay the Prey's Self-Limitation Instead

Moving the delay onto the prey's own self-limitation term instead $\frac{dN}{dt} = D(K - N(t-\tau)) - aN(t)P(t)$ ,predation and conversion now instantaneous, is the classical route to delay-induced instability, the two-species analogue of Hutchinson's single-species delayed logistic equation.

### Working Out the Maths

**Step 1:** The equilibrium is unchanged: at a constant solution the delayed term equals its instantaneous value, so $N^*=m/(ea)$, $P^*=D(K-N^*)/(aN^*)$.

**Step 2: Linearise.** Writing $N=N^*+n$, $P=P^*+p$ and dropping second-order terms:

$$\begin{aligned}
\frac{dn}{dt} &= -\alpha n(t) - \beta p(t) - D n(t-\tau) \\
\frac{dp}{dt} &= \gamma n(t)
\end{aligned}$$

where now $\alpha=aP^*$ (not $D+aP^*$, $D$ now sits on its own as the coefficient of the delayed term), and $\gamma=eaP^*$ carries no delay.

**Step 3: Characteristic equation:**

$$\lambda^2 + \alpha\lambda + \beta\gamma + D\lambda e^{-\lambda\tau} = 0$$

At $\tau=0$ this becomes $\lambda^2+(\alpha+D)\lambda+\beta\gamma=0$, and $\alpha+D=aP^*+D$ is exactly the *old* $\alpha$ from the conversion-delay model, matching Day 24's non-delayed result as it must.

**Step 4: Matching moduli.** Setting $\lambda=i\omega$ gives a quadratic in $u=\omega^2$:

$$u^2 + \left(\alpha^2 - D^2 - 2\beta\gamma\right) u + (\beta\gamma)^2 = 0$$

**Step 5: Sign test.** The product of roots is $(\beta\gamma)^2$, always non-negative as both roots share a sign, and the sum decides which. Requiring both real (discriminant $\geq0$) and positive (sum $>0$) collapses to one condition:

$$\text{Hopf possible} \iff \alpha < D \iff aP^* < D$$

At this project's reference parameters ($D=1,a=1,e=0.5,m=0.3,K=1\to P^*=0.667$): $aP^*=0.667<D=1$ the condition **holds**.

### Confirming It Numerically

```{r selflim-setup}
#| message: false
#| warning: false
#| output: false
predator_prey_rhs_selflim <- function(t, state, parms) {
  N <- max(state[1], 0); P <- max(state[2], 0)
  D <- parms[["D"]]; a <- parms[["a"]]; e <- parms[["e"]]
  m <- parms[["m"]]; K <- parms[["K"]]; tau <- parms[["tau"]]; N0 <- parms[["N0"]]
  lag_N <- if (t <= tau) N0 else max(lagvalue(t - tau, 1), 0)
  dN <- D * (K - lag_N) - a * N * P
  dP <- e * a * N * P - m * P
  list(c(dN, dP))
}

check_hopf_feasibility_selflim <- function(parms = predator_prey_params_delay) {
  D <- parms$D; a <- parms$a; e <- parms$e; m <- parms$m; K <- parms$K
  N_star  <- m / (e * a)
  P_star  <- D * (K - N_star) / (a * N_star)
  data.frame(N_star = N_star, P_star = P_star, alpha_s = a * P_star, D = D,
             hopf_possible = (a * P_star) < D)
}

hopf_check_selflim <- check_hopf_feasibility_selflim()
print(hopf_check_selflim)
cat(sprintf(
  paste0(
    "Self-limitation-delay Hopf check: a*P* = %.4f vs D = %.4f -- %s\n",
    "(derived condition: Hopf possible iff a*P* < D; contrast with the conversion-\n",
    "delay model, where the analogous condition could never hold for any parameters)\n"
  ),
  hopf_check_selflim$alpha_s, hopf_check_selflim$D,
  ifelse(hopf_check_selflim$hopf_possible, "HOPF POSSIBLE", "not possible")
))

run_predator_prey_selflim <- function(tau, state0 = c(N = 1, P = 0.1), t_max = 300,
                                      parms = predator_prey_params_delay) {
  parms$tau <- tau
  parms$N0  <- state0[["N"]]
  out  <- as.data.frame(dede(y = state0, times = seq(0, t_max, by = 0.1),
                             func = predator_prey_rhs_selflim, parms = parms))
  late <- out[out$time > t_max * 0.6, ]
  data.frame(tau = tau, max_N = max(late$N), min_N = min(late$N),
             max_P = max(late$P), min_P = min(late$P),
             rel_amplitude_P = (max(late$P) - min(late$P)) / mean(late$P))
}

tau_seq_selflim  <- c(0, 0.5, 1, 1.5, 2, 2.2, 2.4, 2.6, 2.8, 3, 4, 5, 7, 10, 15, 20)
selflim_sweep_df <- bind_rows(lapply(tau_seq_selflim, run_predator_prey_selflim))
```

```{r selflim-plot}
#| message: false
#| warning: false
ggplot(selflim_sweep_df, aes(x = tau, y = rel_amplitude_P)) +
  geom_line(color = "#C44E52", linewidth = 1) + geom_point(color = "#C44E52", size = 1.5) +
  labs(x = "tau (delay in prey self-limitation)", y = "Relative amplitude of P (settled)",
       title = "Delaying the prey's self-limitation: a genuine Hopf bifurcation appears",
       subtitle = "Same D=1,a=1,e=0.5,m=0.3,K=1 -- analytic condition a*P*=0.667 < D=1 predicts yes") +
  theme_minimal()
```

The onset is sharp, between $\tau=2.2$ (amplitude still $0.02$) and $\tau=2.4$ (amplitude jumps to $0.66$), landing almost exactly on the $\tau\approx2.4$ hand estimate. A persistence check at $\tau=10$ across `t_max = 300, 1500, 3000` gives `rel_amplitude_P = 3.06, 2.95, 2.92`, which is steady, confirming a sustained limit cycle rather than an unsettled transient.

**But there's a real problem.** The `min_N` column shows $N$ going negative right at onset and dramatically more negative as $\tau$ grows, down to $-15.5$ by $\tau=20$. This toy model has no saturating mechanism to cap the oscillation, so past the bifurcation the "population" swings through biologically nonsensical negative values. The Hopf bifurcation itself is real but the model can't be trusted as a faithful biological picture much beyond its onset.

------------------------------------------------------------------------

## A Third Variant: Delay Only the Predator's Intake, Not Its Own Growth

A natural intermediate case: what if only $P(t-\tau)$ is delayed inside the predator's growth term, while $N(t)$ there stays instantaneous?

$$\begin{aligned}
\frac{dN}{dt} &= D(K - N(t)) - aN(t)P(t) \\
\frac{dP}{dt} &= eaN(t)P(t-\tau) - mP(t)
\end{aligned}$$

### The Maths

**Step 1: Linearise.** $\frac{dp}{dt}=\gamma n(t)+mp(t-\tau)-mp(t)$: the $n$ term loses its delay, but $mp(t-\tau)$ survives (it comes from $N^*$, a constant, multiplying the still-delayed $P(t-\tau)$).

**Step 2: Characteristic equation**, with $\alpha=D+aP^*$ as in the original model, but now $\beta\gamma$ carries no $e^{-\lambda\tau}$ factor:

$$(\lambda+\alpha)\left(\lambda+m-me^{-\lambda\tau}\right) + \beta\gamma = 0$$

**Step 3: Match moduli:**

$$u^2 + \left(\alpha^2 - 2\beta\gamma\right) u + \beta\gamma\left(2\alpha m + \beta\gamma\right) = 0$$

The product of roots, $\beta\gamma(2\alpha m+\beta\gamma)$, is always positive ; same situation as the self-limitation model, so both roots share a sign and the sum decides which:

$$\text{Hopf possible} \iff 2\beta\gamma > \alpha^2$$

**Step 4: Reduce to a threshold on D.** Writing $P^*=DC$ ($C=(K-N^*)/(aN^*)$, independent of $D$) reduces this to a genuine threshold on $D$, unlike the original model's absolute impossibility or the self-limitation model's condition (which held outright at the reference parameters):

$$D < \frac{2amC}{(1+aC)^2}$$

At the reference parameters ($N^*=0.6, C=0.667$), that threshold works out to $\sim0.144$. $D=1$ is far above it, so no Hopf is predicted there. This is tested below alongside $D=0.1$, comfortably under the threshold.

### Confirming It Numerically

```{r pdelay-setup}
#| message: false
#| warning: false
#| output: false
predator_prey_rhs_pdelay <- function(t, state, parms) {
  N <- max(state[1], 0); P <- max(state[2], 0)
  D <- parms[["D"]]; a <- parms[["a"]]; e <- parms[["e"]]
  m <- parms[["m"]]; K <- parms[["K"]]; tau <- parms[["tau"]]; P0 <- parms[["P0"]]
  lag_P <- if (t <= tau) P0 else max(lagvalue(t - tau, 2), 0)
  dN <- D * (K - N) - a * N * P
  dP <- e * a * N * lag_P - m * P
  list(c(dN, dP))
}

run_predator_prey_pdelay <- function(tau, state0 = c(N = 1, P = 0.1), t_max = 300,
                                     parms = predator_prey_params_delay) {
  parms$tau <- tau; parms$P0 <- state0[["P"]]
  out  <- as.data.frame(dede(y = state0, times = seq(0, t_max, by = 0.1),
                             func = predator_prey_rhs_pdelay, parms = parms))
  late <- out[out$time > t_max * 0.6, ]
  data.frame(tau = tau, max_N = max(late$N), min_N = min(late$N),
             max_P = max(late$P), min_P = min(late$P),
             rel_amplitude_P = (max(late$P) - min(late$P)) / mean(late$P))
}

params_small_D  <- modifyList(predator_prey_params_delay, list(D = 0.1))
tau_seq_pdelay  <- c(0, 0.5, 1, 2, 3, 5, 7, 10, 15, 20)

pdelay_sweep_refD   <- bind_rows(lapply(tau_seq_pdelay, run_predator_prey_pdelay)) %>%
  mutate(D_label = "D = 1 (above threshold)")
pdelay_sweep_smallD <- bind_rows(lapply(tau_seq_pdelay, run_predator_prey_pdelay,
                                        parms = params_small_D)) %>%
  mutate(D_label = "D = 0.1 (below threshold)")
pdelay_sweep_df <- bind_rows(pdelay_sweep_refD, pdelay_sweep_smallD)
```

```{r pdelay-plot}
#| message: false
#| warning: false
ggplot(pdelay_sweep_df, aes(x = tau, y = rel_amplitude_P, color = D_label)) +
  geom_line(linewidth = 1) + geom_point(size = 1.5) +
  labs(x = "tau (delay in P(t-tau) term only, N instantaneous)", y = "Relative amplitude of P",
       title = "Delaying only P inside the intake term: a threshold on D, not an absolute block",
       subtitle = "Derived threshold D < 0.144 -- tested on both sides", color = NULL) +
  theme_minimal()
```

At $D=1$ (above threshold), `rel_amplitude_P` climbs smoothly from $1.8\times10^{-10}$ to $0.145$ by $\tau=20$, the same non-Hopf shape as the original conversion-delay model. At $D=0.1$ (below threshold, Hopf predicted), the climb is smaller and equally gradual, reaching only $0.033$ by $\tau=20$, with no sharp transition visible.

A persistence check at $D=0.1,\tau=10$ settles it: `rel_amplitude_P` at `t_max = 300, 1500, 3000` is $0.00164\to9.18\times10^{-14}\to1.04\times10^{-15}$ .This is collapsing toward machine-precision zero, the same "critical slowing down" signature as earlier, clearly not a sustained cycle. So even though $D=0.1$ satisfies the derived existence condition, $\tau=10$ isn't past this variant's real onset,so the true critical $\tau$ sits beyond 20, not yet located. The threshold on $D$ is correctly derived, but pinning down *where* the bifurcation occurs for a given $D$ needs a wider sweep.

------------------------------------------------------------------------

## What This All Adds Up To

Three loose ends from Day 25 closed cleanly: $\alpha$'s transition window is now a precise $(0.105, 0.178)$ bracket, the $\tau=40/80$ amplitude climb is confirmed to decay to machine-precision zero, and the $\gamma=305.8$ dip turned out to be neither "real" nor "transient" - its non-monotonic flipping points at an unsettled oscillation at the sweep's own operating point, casting doubt on $\gamma$ and $\kappa$'s "modest"/"flat" Day 25 conclusions.

The deeper result is about *where* a delay sits, not whether delay can destabilise a system in general. Delaying the predator's conversion (Day 25's model) can never produce a Hopf bifurcation, proven by factoring the destabilising quantity into $-mD(K+N^*)/N^*$, a bare positive $D$ multiplying an unconditionally negative bracket. Delaying the prey's self-limitation instead does produce one, confirmed both analytically ($aP^*<D$) and by simulation landing the onset almost exactly where predicted. Though the limit cycle's amplitude grows unboundedly past onset, driving prey biomass negative, a limitation of this toy model rather than the underlying mechanism. A third, intermediate variant (delay only the predator's intake) derives a genuine threshold on $D$ rather than an absolute block, but the tested $\tau$ wasn't far enough to actually see it.

------------------------------------------------------------------------

## What's Next

- **Find the actual critical tau for the P-only-delay variant at D=0.1:** Extend `tau_seq_pdelay` well past 20 (e.g. up to 40–60) and rerun the persistence check at each candidate onset to locate where `rel_amplitude_P` stops decaying and starts holding steady.
- **Characterise how far past onset the self-limitation model's amplitude stays biologically sensible:** `min_N` is already negative at $\tau=2.4$ (barely past onset) and reaches $-15.5$ by $\tau=20$. It's worth finding the $\tau$ range where $N$ stays non-negative, and considering whether a saturating term (a proper functional response, rather than mass-action) would cap the amplitude the way it should.
- **Check directly whether `capacity_mult=10, resource_decrease=0.001` is actually oscillating:** Rather than reading single-timestep snapshots at different `t_max`. A real time series at this exact operating point would settle whether the gamma/kappa sweep's noise floor is genuine unsettled oscillation or something else.
- **Rerun gamma and kappa's isolation sweeps:** Do this with a windowed-average read (the same "late window" convention used throughout the delay-model work) instead of a single final-timestep snapshot, to check whether their "modest"/"flat" Day 25 conclusions survive averaging out whatever phase-sampling noise today's gamma recheck exposed.
- **Carry these toy-model findings back over to the actual size-spectrum (mizer) model:** Everything above was worked out on the abstract two-species predator-prey system, not on the anchovy model itself - the natural next step is checking which of these delay mechanisms, if any, has a real biological analogue in the size-spectrum setting, and whether it could plausibly help explain the juvenile-pileup dynamics this project has been tracking since Day 6.
