---
title: "Day 21: Watching Hysteresis Play Out in Time, Not Just in the Bifurcation Diagram"
author: "Samia"
date: "2026-07-06"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, limit-cycle, hysteresis, time-series, diffusion, blog]
description: "Day 20's headline result was corrected first: the wider sweep doesn't close the forward/backward gap, it opens a much bigger one. Today follows that up by watching biomass directly over time instead of only the settled max/min, at a handful of resource_decrease values, run first independently and then as a proper chained forward/backward pair. The chained version shows the hysteresis directly: at resource_decrease = 0.05, forward finds a healthy steady population while backward is still collapsed from a higher value. A few further threads (Beverton-Holt recruitment, a reordered forward/backward chain, a single-point diffusion check) were started but are parked unfinished, with known issues still in the script."
---

## Where Day 20 Left Us

Day 20 widened Day 18's narrow, ambiguous 10-point sweep to a 40-point, log-spaced grid and reran it on mizer 3.1's second-order finite-volume scheme plus external diffusion. The post originally read that result backwards: it said the forward/backward gap closed. Re-reading the actual figure showed the opposite — the backward branch collapses to zero around `resource_decrease ≈ 0.02`, while the forward branch keeps a nonzero cycle out to `≈ 0.15–0.2`, a wide and unambiguous hysteresis loop, much bigger than Day 18's narrow `rd ≈ 0.008` ambiguity. The title, description, and body of that post have all been corrected to say so. The diffusion caveat stands either way: `ext_diffusion` was filled as a flat constant across all sizes rather than mizer's own allometric `w^(n+1)` shape, so the result (now: a wide hysteresis loop, not a closed gap) still needs to be reproduced with diffusion set up properly before it can be trusted.

Today picks that fix up directly, and then goes further: instead of only ever looking at each `resource_decrease` value's settled max/min biomass, it looks at biomass over time itself, to see the shape of the oscillation (or its absence) directly, and to watch the hysteresis play out step by step rather than reading it off a summary curve.

```{r}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse)
library(glue)
library(ggplot2)
library(colorRamps)
library(future.apply)
```

------------------------------------------------------------------------

## Setup

`make_second_order_params()` carries over the second-order finite-volume scheme from Day 20, but fixes the diffusion setup that post's "A Catch" section flagged as wrong. Instead of filling an `ext_diffusion` array with a flat constant, it now sets the baseline `D_ext` rate directly in `species_params` via `given_species_params()`, and lets mizer derive its own `w^(n+1)` allometric shape from that rate:

```{r}
#| message: false
#| warning: false
#| output: false
make_second_order_params <- function(lambda = 2.05, resource_decrease = 0.001,
                                     second_order = TRUE, ext_diff = 0.01) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  params <- newSingleSpeciesParams(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )
  r      <- getResourceRate(params) * resource_decrease
  params <- setResource(params, resource_rate = r,
                        resource_dynamics = "resource_semichemostat",
                        balance = TRUE)

  # given_species_params<-() is mizer's "stronger" setter: unlike
  # species_params<-(), it also triggers a recalculation of the derived
  # parameters, so setting the baseline D_ext rate here and letting mizer
  # recompute is the fix Day 20 flagged as still needed -- editing the
  # baseline rate and letting setExtDiffusion()'s w^(n+1) shape come from
  # mizer itself, rather than constant-filling the array by hand.
  if (second_order) {
    second_order_w(params) <- c(flux = "centred", bin_average = TRUE)
  }
  given_species_params(params)$D_ext <- ext_diff
  params
}
```

`rd_time_series` picks five `resource_decrease` values spanning Day 20's hysteresis region: deep in the oscillating regime, through the transition, and out past where the backward branch had collapsed.

```{r}
#| message: false
#| warning: false
#| output: false
rd_time_series <- c(0.001, 0.01, 0.05, 0.15, 0.3)
```

------------------------------------------------------------------------

## Experiment: Biomass Over Time, Run Independently

The first pass runs each `resource_decrease` value on its own, 300 years, starting from mizer's default initial state every time rather than carrying anything over between runs:

```{r}
#| message: false
#| warning: false
#| output: false
biomass_over_time <- lapply(rd_time_series, function(rd) {
  p   <- make_second_order_params(resource_decrease = rd)
  sim <- project(p, t_max = 300, dt = 0.1, t_save = 0.5,
                progress_bar = FALSE, effort = 0, method = "tr_bdf2")
  bm  <- getBiomass(sim)[, "Anchovy"]
  data.frame(time = as.numeric(names(bm)), biomass = as.numeric(bm),
            resource_decrease = factor(rd))
})
biomass_over_time_df <- bind_rows(biomass_over_time)
```

```{r}
#| message: false
#| warning: false
ggplot(biomass_over_time_df, aes(x = time, y = biomass, color = resource_decrease)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time", y = "Biomass",
       title = "Biomass over time for a few resource_decrease values",
       subtitle = "Flat line = settled fixed point; oscillating = limit cycle",
       color = "resource_decrease") +
  theme_minimal()
```

This is the baseline case: each value is only ever approached one way, from the same default starting point, so it says nothing yet about direction-dependence. It's the "normal" run the next section compares against.

------------------------------------------------------------------------

## Experiment: The Same Values, Chained Forward and Backward

The more useful version carries state between steps the same way Day 20's `fwd_so`/`bwd_so` sweep did: each `resource_decrease` value's run starts from the *previous* value's settled abundances, instead of resetting every time. `run_biomass_time_series()` does that, keeping the full biomass trace (not just the settled max/min) for every step, with time offset cumulatively across steps so each direction reads as one continuous trajectory:

```{r}
#| message: false
#| warning: false
#| output: false
run_biomass_time_series <- function(rd_seq, init_n = NULL, init_n_pp = NULL, t_run = 300) {
  state_n   <- init_n
  state_npp <- init_n_pp
  t_offset  <- 0
  segments  <- vector("list", length(rd_seq))

  for (i in seq_along(rd_seq)) {
    p <- make_second_order_params(resource_decrease = rd_seq[i])
    if (!is.null(state_n)) {
      p@initial_n[]    <- state_n
      p@initial_n_pp[] <- state_npp
    }
    sim <- project(p, t_max = t_run, dt = 0.1, t_save = 0.5,
                   progress_bar = FALSE, effort = 0, method = "tr_bdf2")
    bm   <- getBiomass(sim)[, "Anchovy"]
    tv   <- as.numeric(names(bm))
    last <- dim(sim@n)[1]

    state_n   <- sim@n[last, , ]
    state_npp <- sim@n_pp[last, ]

    segments[[i]] <- data.frame(time = tv + t_offset, biomass = as.numeric(bm),
                                resource_decrease = rd_seq[i])
    t_offset <- t_offset + max(tv)
  }
  list(df = bind_rows(segments), n_final = state_n, npp_final = state_npp)
}

fwd_ts <- run_biomass_time_series(rd_time_series)
bwd_ts <- run_biomass_time_series(rev(rd_time_series),
                                  init_n    = fwd_ts$n_final,
                                  init_n_pp = fwd_ts$npp_final)

biomass_fwd_bwd_df <- bind_rows(
  fwd_ts$df %>% mutate(direction = "Forward"),
  bwd_ts$df %>% mutate(direction = "Backward")
)
```

```{r}
#| message: false
#| warning: false
ggplot(biomass_fwd_bwd_df, aes(x = time, y = biomass, color = direction)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (cumulative across steps)", y = "Biomass",
       title = "Biomass over time: forward vs backward through the same resource_decrease values",
       subtitle = "Each segment holds resource_decrease fixed for 300 years before stepping to the next value",
       color = "Direction") +
  theme_minimal()
```

### What This Shows

Reading the five 300-year segments left to right:

**Forward** (`0.001 → 0.01 → 0.05 → 0.15 → 0.3`): strong limit-cycle oscillation at `rd = 0.001` (biomass swinging roughly `0.02–0.22`), settling almost immediately to a steady `~0.085` at `rd = 0.01`, then a lower steady `~0.055` at `rd = 0.05`, then collapsing to near-zero for both remaining segments (`rd = 0.15`, `0.3`).

**Backward** (`0.3 → 0.15 → 0.05 → 0.01 → 0.001`, starting from wherever forward ended up, already collapsed): stays collapsed near zero through *all three* of `rd = 0.3`, `0.15`, and `0.05`, only recovering to the same `~0.085` steady state once `rd` drops to `0.01`, and only re-entering the large oscillation once `rd` reaches `0.001`.

The disagreement at `rd = 0.05` is the headline: forward finds a healthy `~0.055` steady population there, backward is still fully collapsed. Which state the model is in at that `resource_decrease` value depends on which direction it was approached from, not on the value itself. That's the same wide hysteresis loop Day 20 found in the bifurcation diagram, now visible directly in the time series rather than inferred from a settled max/min summary.

------------------------------------------------------------------------

## Parked, Unfinished Threads

A few further ideas got started in `21_experiments.R` today but aren't in a state to draw conclusions from yet, and are left here as flagged rather than fixed:

- **Reordering the forward/backward chain:** a draft ran the backward pass first (fresh, no carried-over state) and chained forward from backward's ending, the reverse of the convention used above and in Day 20. Whether that ordering matters for the shape of the loop hasn't been checked, since it hasn't been re-run cleanly since the swap.
- **Wiring in `setBevertonHolt()` recruitment:** an attempt to add stock-recruitment density dependence into `make_second_order_params()` currently builds the `params` object twice (the first build, including the `given_species_params()$D_ext` assignment, gets discarded before `setBevertonHolt()` runs on a second, fresh build), so as written it isn't actually testing what it looks like it's testing.
- **A single-point check at `resource_decrease = 0.5`** using `plotHover()` for an interactive look at biomass, referencing `lambda` and `second_order` as loose variables that are never assigned anywhere in the script itself — it only runs if those happen to still be sitting in the R session from something run by hand.
- There's also a stray syntax error (`run_rd_sweep_second_order(rd_seq_so,,` — a double comma) sitting in the sweep block above these three, which would need fixing before the script can be sourced top-to-bottom at all.

------------------------------------------------------------------------

## What's Next

- **Fix the script issues above** before picking any of the three parked threads back up: the syntax error blocking a clean source, the double `params` build in front of `setBevertonHolt()`, and the loose `lambda`/`second_order` variables in the `resource_decrease = 0.5` check.

- **Settle on one forward/backward convention** for future chained comparisons, rather than swapping which direction runs fresh vs chained ad hoc between drafts.

- **Finish the Beverton-Holt attempt properly** once the double-build is fixed, to see whether stock-recruitment density dependence changes the size of the hysteresis loop found on Day 20/21.

- **Extend the time-series overlay beyond five points** — ideally the same 40-point log-spaced grid from Day 20 — to see whether the `rd = 0.05` disagreement found today is an isolated point or part of a wider direction-dependent stretch.
