---
title: "Day 18: The Limit Cycle Is a Global Attractor, but Supercritical vs Subcritical Is Still Open"
author: "Samia"
date: "2026-07-01"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, limit-cycle, stability, perturbation, bifurcation, hysteresis, hopf, blog]
description: "Two experiments today. First: five perturbation magnitudes spanning three orders of magnitude are applied to the mature stock; all converge to the same oscillatory orbit by t = 550, confirming the limit cycle is a stable global attractor. Second: a forward-backward sweep through resource_decrease with diffusion switched off tests for hysteresis. Amplitude grows continuously from zero on the forward sweep, a supercritical signature, but the backward sweep sustains oscillation to a noticeably higher resource_decrease than the forward onset — a gap too large to write off as a sweep-start artefact alone, so a purely supercritical Hopf bifurcation can't yet be confirmed."
---

## Where Day 17 Left Us

Day 17 identified a weakness in the initialisation procedure used throughout the preceding experiments. The two-stage perturbation in `make_limit_cycle_sim`, depleting the resource to destabilise the system, then knocking down mature fish by a fixed factor of 1000, was the mechanism used to kick the system off the unstable fixed point and onto the limit cycle. But this is an ad-hoc procedure. It does not tell us whether the limit cycle is a genuine attractor (i.e. trajectories from *arbitrary* initial conditions converge to it) or whether it is fragile, reached only by this particular path.

Day 17's "What's Next" proposed a direct test: vary the perturbation magnitude across a wide range and check whether all trajectories converge to the same orbit. If they do, the limit cycle is a stable attractor and its basin of attraction is at least as large as the range of perturbations tested. If they diverge, the initialisation is doing more work than assumed.

Today runs exactly that experiment.

```{r}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse)
library(glue)
library(ggplot2)
library(colorRamps)
```

------------------------------------------------------------------------

## Setup

The helper functions are carried over from previous days. The one change from Day 17 is that `make_limit_cycle_sim` now accepts a `perturbation` argument (the factor by which the mature stock is divided at $t = 10$) and uses the `tr_bdf2` integrator for improved stability.

```{r}
#| message: false
#| warning: false
#| output: false
make_params <- function(lambda = 2.05, resource_decrease = 0.001) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  params <- newSingleSpeciesParams(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )
  r      <- getResourceRate(params) * resource_decrease
  params <- setResource(params, resource_rate = r,
                        resource_dynamics = "resource_semichemostat")
  params
}

make_limit_cycle_sim <- function(params, t_total = 600, effort = 0, perturbation = 1e-3) {
  params@initial_n_pp[] <- params@cc_pp * 0.1

  sim_init <- project(params, t_max = 10, dt = 0.1, t_save = 0.2,
                      progress_bar = FALSE, effort = 0,
                      method = "tr_bdf2")
  idx  <- params@w >= 10 & params@w <= 100
  last <- dim(sim_init@n)[1]
  sim_init@n[last, , idx] <- sim_init@n[last, , idx] / perturbation

  project(sim_init, t_max = t_total - 10, dt = 0.1, t_save = 0.2,
          progress_bar = FALSE, effort = effort,
          method = "tr_bdf2")
}

test_perturbation <- function(perturbation, lambda = 2.05, t_total = 600) {
  p   <- make_params(lambda = lambda)
  sim <- make_limit_cycle_sim(p, t_total = t_total, perturbation = perturbation)
  bm  <- getBiomass(sim)[, "Anchovy"]
  t   <- as.numeric(names(bm))
  list(sim = sim, bm = bm, t = t)
}
```

------------------------------------------------------------------------

## Experiment: Varying the Perturbation Magnitude

Five perturbation levels are tested: 1, 5, 10, 100, and 1000. A perturbation of 1000 means the mature stock ($w \geq w_\text{mat}$) is divided by 1000 at $t = 10$, leaving it at 0.1% of its value just before the knock-down. A perturbation of 1 means no knock-down is applied at all as the system evolves from the depleted-resource initial condition alone.

```{r}
#| message: false
#| warning: false
#| output: false
perturbation    <- c(1e3, 1e2, 1e1, 5, 1)
results_fishing <- lapply(perturbation, test_perturbation)
```

------------------------------------------------------------------------

## Late-Time Overlay ($t > 550$)

To make the convergence precise, the trajectories are overlaid for $t > 550$ only. By this point any transient from the initial condition should have decayed, leaving only the attractor.

```{r}
#| message: false
#| warning: false
plot(NULL,
     xlim = c(550, 600),
     ylim = range(sapply(results_fishing, function(r) range(r$bm[r$t > 550]))),
     xlab = "Time (years)", ylab = "Biomass",
     main = "Biomass trajectories (t > 550)")
cols <- colorRampPalette(c("blue", "red"))(length(perturbation))
for (i in seq_along(results_fishing)) {
  r    <- results_fishing[[i]]
  mask <- r$t > 550
  lines(r$t[mask], r$bm[mask], col = cols[i])
}
legend("topright",
       legend = paste0("Perturb = ", perturbation),
       col = cols, lty = 1, cex = 0.8)
```

------------------------------------------------------------------------

## What the Plot Shows

The five trajectories are essentially indistinguishable after $t = 550$. They share the same amplitude (oscillating between roughly $0.03$ and $0.16$), the same period (approximately 7–8 years), and the same waveform shape. The only visible residual difference is a slight phase offset between some lines, most visible around the peaks near $t \approx 570$–$575$, but even this appears to be shrinking over the interval shown.

This is direct evidence for two properties of the system:

**The fixed point is unstable:** None of the trajectories converge to a constant biomass. Every initial condition, including the case with no knock-down (Perturbation = 1), evolves into sustained oscillation. The fixed point is present in the dynamics (the trajectories oscillate around it) but it does not attract any of the tested initial conditions.

**The limit cycle is a stable attractor:** All five trajectories, starting from initial conditions that differ by three orders of magnitude in the mature stock, converge to the same periodic orbit. This is the defining property of a stable (attracting) limit cycle: it draws in trajectories from its neighbourhood and, as the experiment shows, from a very wide neighbourhood. The basin of attraction extends at least from Perturbation = 1 to Perturbation = 1000, which covers the full range of plausible starting states for this system.

**The basin of attraction is large:** Perturbation = 1000 corresponds to reducing the adult stock to 0.1% of its value which is an extreme disturbance that would, in a stable-equilibrium system, either drive the population to extinction or require a very long recovery transient. Here it does neither: the system returns to the same orbit on a timescale comparable to the perturbation = 1 case. The cohort-wave mechanism is robust enough to self-organise from a near-empty adult population.

------------------------------------------------------------------------

## What This Means for the Initialisation Procedure

The experiment also validates the two-stage initialisation used throughout Days 10–17. The specific choice of dividing the mature stock by 1000 was arbitrary; the results here confirm that essentially any perturbation large enough to break the system away from the vicinity of the fixed point produces the same long-run behaviour. The initialisation procedure is not injecting a particular orbit; it is simply providing a push that allows the natural attractor to assert itself.

------------------------------------------------------------------------

## Hysteresis Test: Is the Bifurcation Subcritical or Supercritical?

Day 17 proposed a second diagnostic: sweep the `resource_decrease` parameter forward (stable → oscillating) and then backward (oscillating → stable), carrying the final state of each run into the next. If the system shows hysteresis, the limit cycle persisting at lower `resource_decrease` on the way back than where it appeared on the way forward,then that is evidence of a subcritical Hopf bifurcation, with a bistable region where both the fixed point and the limit cycle are stable simultaneously. If the two sweeps coincide, the bifurcation is supercritical.

Mizer includes a diffusion term arising from variability in prey sizes (the jump-growth equation). Before attempting to switch it off, it is worth checking whether it is actually non-zero for this parameterisation.

```{r}
#| message: false
#| warning: false
p_check <- make_params(lambda = 2.05)
range(getDiffusion(p_check))
```

`getDiffusion` returns zero across all size classes. The diffusion term is already zero for this single-species anchovy parameterisation and the jump-growth diffusion evaluates to zero because the effective variance in prey sizes collapses to zero under this resource and feeding-kernel setup. No special handling is needed; `make_params` is used directly and the model is already in the pure advection-mortality regime.

The sweep helper runs each `resource_decrease` value for 300 years, uses the last 40% of each run to estimate oscillation amplitude, and passes the final state directly into the next run's initial condition.

```{r}
#| message: false
#| warning: false
#| output: false
run_rd_sweep <- function(rd_seq, init_n = NULL, init_n_pp = NULL,
                         t_run = 300, lambda = 2.05, label = "") {
  out       <- data.frame(rd = rd_seq, amp = NA_real_, bm_mean = NA_real_)
  state_n   <- init_n
  state_npp <- init_n_pp

  for (i in seq_along(rd_seq)) {
    p <- make_params(lambda = lambda, resource_decrease = rd_seq[i])
    if (!is.null(state_n)) {
      p@initial_n[]    <- state_n
      p@initial_n_pp[] <- state_npp
    }
    sim <- project(p, t_max = t_run, dt = 0.1, t_save = 0.5,
                   progress_bar = FALSE, effort = 0, method = "tr_bdf2")
    bm   <- getBiomass(sim)[, "Anchovy"]
    tv   <- as.numeric(names(bm))
    late <- bm[tv > t_run * 0.6]
    last <- dim(sim@n)[1]

    state_n   <- sim@n[last, , ]
    state_npp <- sim@n_pp[last, ]

    out$amp[i]     <- max(late) - min(late)
    out$bm_mean[i] <- mean(late)
  }
  list(df = out, n_final = state_n, npp_final = state_npp)
}

rd_seq <- seq(0.001, 0.01, length.out = 10)

fwd <- run_rd_sweep(rd_seq, label = "FWD")
bwd <- run_rd_sweep(rev(rd_seq),
                    init_n    = fwd$n_final,
                    init_n_pp = fwd$npp_final,
                    label     = "BWD")
```

```{r}
#| message: false
#| warning: false
fwd_df <- fwd$df
bwd_df <- bwd$df[order(bwd$df$rd), ]

thresh     <- 0.005
fwd_onset  <- fwd_df$rd[which(fwd_df$amp  > thresh)[1]]
bwd_offset <- bwd_df$rd[tail(which(bwd_df$amp > thresh), 1)]

plot(fwd_df$rd, fwd_df$amp,
     type = "l", col = "steelblue", lwd = 2,
     xlab = "resource_decrease",
     ylab = "Oscillation amplitude (max − min biomass)",
     main = "Hysteresis test",
     ylim = c(0, max(fwd_df$amp, bwd_df$amp) * 1.1))
lines(bwd_df$rd, bwd_df$amp, col = "firebrick", lwd = 2)
abline(v = fwd_onset,  lty = 2, col = "steelblue")
abline(v = bwd_offset, lty = 2, col = "firebrick")
legend("topleft",
       legend = c(
         sprintf("Forward  — onset  rd ≈ %.4f", fwd_onset),
         sprintf("Backward — offset rd ≈ %.4f", bwd_offset)
       ),
       col = c("steelblue", "firebrick"), lwd = 2)
```

------------------------------------------------------------------------

## What the Hysteresis Plot Shows

The forward and backward curves both drop toward zero amplitude in roughly the same region, and neither jumps discontinuously — encouraging signs for a supercritical bifurcation. But they don't fully coincide either. The backward curve keeps oscillating down to $\text{rd} \approx 0.008$, while the forward curve's amplitude has already fallen toward the noise floor well before that point. Read at face value, that gap is exactly what a subcritical bifurcation with a small bistability window would look like: the limit cycle outliving its own forward onset on the way back down.

The obvious objection is that the forward "onset" detector is contaminated: it's flagged at rd = 0.001 simply because the sweep starts there, already inside the oscillating regime. That's true, but it only explains away half the gap — it means the forward transition point isn't well measured by `fwd_onset`, not that the gap itself isn't real. The backward offset at $\text{rd} \approx 0.008$ is a direct, uncontaminated measurement, and the forward curve's amplitude is visibly near zero below that value on the plot. Critical slowing down near the bifurcation is a real confound too: a 300-year run may not fully settle before the next step is taken, which would inflate the backward curve's apparent transition point without any true hysteresis behind it. But at this sweep resolution and run length, that explanation can't be distinguished from genuine bistability.

So the two signatures don't agree. Continuous, square-root-like amplitude growth on the forward sweep ($A \propto \sqrt{|r - r_c|}$, no discontinuous jump) is a supercritical signature. The unresolved gap between the forward and backward transition points is what a subcritical bifurcation with a bistable window would produce. **Given that, this data can't yet confirm the bifurcation is supercritical.** It may be subcritical with a narrow bistable region that a coarse 10-point sweep and 300-year runs aren't resolving cleanly, or it may be supercritical with the whole gap explained by critical slowing down. Distinguishing the two needs either much longer per-step runs (to let slow convergence fully resolve) or a finer `rd` grid around 0.008 to see whether the gap holds up or recedes. That's left open rather than settled here.

The effort scans in Days 16–17 are sometimes read as a second confirmation of the same picture: the transition back to a stable fixed point at high fishing effort also showed damped spiralling rather than a discontinuous jump. But calling that "consistent with supercritical" is a weaker claim than it sounds, and it's now doubly so given the hysteresis gap above. The spiral there persists for a good number of cycles before settling, longer than critical slowing down alone would predict, and it isn't a numerical-integration artefact either, since the same solver and timestep produce clean, non-decaying oscillation everywhere else in this post. Asserting consistency glosses over that persistence rather than explaining it. Rather than a blanket statement, this needs a closer look: plotting the maximum and minimum of each cycle separately as effort increases (see What's Next) rather than the collapsed max − min amplitude used so far.

------------------------------------------------------------------------

## What's Next

- **Try a different perturbation type** to see whether the same convergence-to-attractor result holds no matter what form the perturbation takes, rather than relying solely on the mature-stock knock-down used so far.

- **Run with `balance = FALSE` in setResource.** The effects should be more realistic, since the model is not inflating some rates to compensate for others.

- **Plot maximum and minimum separately** instead of plotting max − min, to get a bifurcation-diagram look at the amplitude data rather than a single collapsed amplitude curve.

- **Resolve whether the forward/backward gap at `rd ≈ 0.008` is real hysteresis or a run-length artefact.** Rerun the sweep with much longer per-step runs (to let critical slowing down fully settle) and a finer `resource_decrease` grid around 0.008, since the current 10-point, 300-year sweep can't distinguish a genuinely subcritical bifurcation with a narrow bistable window from a supercritical one confounded by slow convergence.
