---
title: "Day 7: What's Driving the Oscillation?"
author: "Samia"
date: "2026-06-16"
categories: [research, mizer, oscillations, feeding-level, bifurcation, blog]
description: "Chasing the mechanism behind the oscillation going from a dead end with EGrowth animations to a clear result about food limitation in the juvenile size classes."
---

## The Question

Day 6 ended with a confirmed limit cycle and a rough period estimate of 7–9 years, close to three generation times. The natural next question: *what's actually driving it?* Two candidate mechanisms were on the table:

- **Food-limitation cycle**: The resource oscillates, fish growth rates follow, maturation is delayed and advanced on alternating half-cycles, and the period is set by that delay.
- **Cohort wave**: A pulse of recruits moves through the size spectrum together, grazing the resource as it passes through small sizes, but individual growth rates barely change. The oscillation is in population structure, not food intake.

The way to distinguish them is to check whether growth rates actually change during the cycle.

------------------------------------------------------------------------

## Setup

Parameters, model construction, and the initial perturbation that starts the limit-cycle transient:

```{r}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse) 
library(glue)
library(ggplot2)

p <- list(
  dt     = 0.001,
  dx     = 0.1,
  w_min  = 0.0003,
  w_max  = 66.5,
  w_mat  = 10,
  alpha  = 0.1,
  gamma  = 750,
  lambda = 2,
  a0     = 100
)

kappa <- p$a0 * exp(-6.9 * (p$lambda - 1))
no_w  <- round(log(p$w_max / p$w_min) / p$dx)

params <- newSingleSpeciesParams(
  species_name = "Anchovy",
  w_min  = p$w_min,  w_max  = p$w_max,  w_mat  = p$w_mat,
  no_w   = no_w,     lambda = p$lambda,  kappa  = kappa,
  alpha  = p$alpha,  gamma  = p$gamma,   ks     = 0
)

r <- resource_rate(params) * 0.001
params <- setResource(params, resource_rate = r,
                      resource_dynamics = "resource_semichemostat")
params@initial_n_pp[] <- params@cc_pp * 0.1

# Spin up, then crash the large size class (10–100 g) by 10^3
sim <- project(params, t_max = 10, dt = 0.1, t_save = 0.2,
               progress_bar = FALSE, method = "predictor-corrector")

rng  <- c(10, 100)
idx  <- params@w >= rng[1] & params@w <= rng[2]
last <- dim(sim@n)[1]
sim@n[last, , idx] <- sim@n[last, , idx] / 10^3

sim <- project(sim, t_max = 90, dt = 0.1, t_save = 0.2,
               progress_bar = FALSE, method = "predictor-corrector")

sim_300 <- project(sim, t_max = 200, dt = 0.1, t_save = 0.2,
                   progress_bar = FALSE, method = "predictor-corrector")#can use tr_bdf2,but needs much finer dt

sim_600 <-  project(sim_300, t_max = 300, dt = 0.1, t_save = 0.2,
                    progress_bar = FALSE, method = "predictor-corrector")#can use tr_bdf2,but needs much finer dt

```

------------------------------------------------------------------------

## Animating EGrowth

The first attempt was to animate `getEGrowth()` across the oscillation — if growth rates are cycling, the curve should visibly rise and fall with the plankton. The mizer `animate()` function handles this directly once you have a long-enough simulation object:

```{r}
#| message: false
#| warning: false
t_trough <- 507.2
t_peak   <- 570.2
animate(getEGrowth(sim_600, time_range = c(t_trough - 1, t_peak + 1)), log = "xy")
```

The animation ran but the curve barely moved. Comparing two specific time points confirmed it. The EGrowth curves at `t = 557` and `t = 559` were essentially identical on a log-log scale, with only a hairline split visible at the very smallest sizes.

------------------------------------------------------------------------

## A Cleaner Comparison

The time points chosen (557 and 559, two years apart) were too close together to be meaningful for a 7–9 year cycle. Peak and trough of the plankton are roughly 3.5–4.5 years apart, so both points were likely at similar phases of the oscillation.

The right comparison is between the actual biomass peak and trough. `plotHover()` on the settled portion of the biomass time series makes it easy to read off the exact times interactively:

```{r}
#| message: false
#| warning: false
plotHover(getBiomass(sim_600), tlim = c(500, 580))

t_trough <- 507.2
t_peak   <- 570.2
```

Then overlaying EGrowth, the size spectrum, and individual spectra at those two times:

```{r}
#| message: false
#| warning: false

plot2(getEGrowth(sim_600, t = t_peak),
      getEGrowth(sim_600, t = t_trough),
      log = "xy")

```

Still essentially identical. The curves sat on top of each other across the full size range. If growth rates don't change, the mechanism has to be somewhere else.

------------------------------------------------------------------------

## Feeding Level: The Right Diagnostic

`getEGrowth` tells you energy available for growth, but it folds in the maturity and metabolic costs. A cleaner measure of whether fish are food-limited is the feeding level $f(w)$ , the fraction of maximum intake rate actually achieved, running from 0 (no food) to 1 (fully saturated).

```{r}
#| message: false
#| warning: false
getFeedingLevel(getParams(sim_600, t_peak))
# Feeding level (1 species x 123 sizes)
#   Anchovy: min=0.896 mean=0.949 max=0.995

getFeedingLevel(getParams(sim_600, t_trough))
# Feeding level (1 species x 123 sizes)
#   Anchovy: min=0.244 mean=0.776 max=0.923
```

This is a large difference. At the plankton peak, fish are feeding at 90–100% of their maximum rate across all size classes meaning they are effectively saturated, and there is no food limitation. At the plankton trough, the smallest fish drop to a feeding level of 0.244 so eating at roughly a quarter of their maximum rate. That's a substantial slowdown in growth for juveniles.

The earlier EGrowth animation showed no difference because the effect is concentrated in the smallest size classes, where the log-log scale compresses things visually. The feeding level comparison makes it explicit.

------------------------------------------------------------------------

## Animating the Feeding Level

With the static peak/trough comparison confirming food limitation, the next step was to animate `getFeedingLevel()` which allows us to watch the feeding level evolve at each size class over time rather than just seeing summary statistics:

```{r}
#| message: false
#| warning: false
animate(getFeedingLevel(sim_600, time_range = c(550, 600)))
```

The result was striking, and raises more questions than it answers.

The oscillation in feeding level is clearly periodic, which confirms the food-limitation cycle. But the pattern across size classes is uneven in a way that wasn't expected:

- **Only the smaller size classes oscillate meaningfully.** Larger fish maintain a near-constant, high feeding level throughout the cycle. Their food supply, larger prey, barely fluctuates.
- **The temporal shape is asymmetric.** The smallest juveniles spend the majority of the cycle at a *high* feeding level, with only a brief, sharp crash before recovering. It is not a smooth sine-wave oscillation but instead looks more like the feeding level sustains near-saturation for an extended period, then collapses rapidly, then climbs back up again. The crash is fast; the recovery and plateau are slow.

This asymmetry is unexpected. In a simple predator-prey model you would expect roughly equal time above and below the mean. The fact that small fish are food-saturated for far longer than they are food-limited suggests the resource recovery is relatively fast compared to the grazing pressure that drives the depletion. Once a grazing cohort passes through the juvenile size range, the plankton bounces back quickly but the crash, while brief, is sharp enough to leave a lasting imprint on cohort growth.

This pattern needs more investigation. It may be that the plankton has a fast intrinsic recovery rate relative to the consumption rate, meaning depletions are short-lived but intense. Alternatively, the size-specificity of the crash could reflect something about how the cohort moves through the size spectrum.

------------------------------------------------------------------------

## What This Means for the Mechanism

The oscillation is food-limitation driven, concentrated in the juvenile size range:

1.  A cohort of juveniles grazes the plankton hard as it passes through small sizes, driving plankton to its trough.
2.  The next wave of recruits enters a depleted food environment, feeding level drops to \~0.24 at the smallest sizes and grows more slowly.
3.  Slower growth means the cohort takes longer to reach maturity, delaying the next grazing pulse.
4.  With fewer juveniles competing for it, the plankton recovers.
5.  The following cohort grows fast on the recovered resource, matures quickly, grazes hard.
6.  The cycle repeats.

The three-generation period from Day 6 is consistent with this: it takes roughly one generation for the depleted-food cohort to reach maturity, one for the recovery to propagate back into reproduction, and one more for that effect to produce a new grazing pulse.

------------------------------------------------------------------------

## Extending the Perturbation

With the mechanism clearer, the perturbation code was extended to also crash the plankton in the prey size range of the large fish, the part of the resource spectrum they feed on most heavily:

```{r}
#| message: false
#| warning: false
#| output: false
beta <- params@species_params$beta

prey_rng <- c(10, 100) / beta
pp_idx   <- params@w_full >= prey_rng[1] & params@w_full <= prey_rng[2]

sim@n_pp[last, pp_idx] <- sim@n_pp[last, pp_idx] / 10^3
```

The result is that the overall perturbation shape is essentially the same. The one visible difference is in the growth rate comparison plot: with the plankton also crashed, the second curve sits slightly *above* the first, whereas without the plankton crash it is the other way round. This makes sense: crashing the plankton directly depresses the feeding level of the remaining small fish immediately after the perturbation, so the first time point reflects a more food-limited state than the second, flipping the relative ordering of the curves. The system still converges to the same limit cycle; only the phase of the transient shifts slightly.

One surprising finding here: changing the magnitude of the initial large fish crash, whether you reduce abundance by 10,10\^3,10\^7 or another factor makes essentially no difference to the subsequent growth rate or the trajectory back toward the limit cycle. The system converges at the same rate regardless. This is strange. Naively, crashing large fish harder should produce a larger transient, and a larger transient might take a different path through the phase space. The fact that it doesn't suggests the system is being pulled very strongly toward the same attractor along the same trajectory, and the initial condition is largely irrelevant once you are far enough from the limit cycle. It may also reflect that the large fish abundance is not the limiting factor in the dynamics. The plankton and the juvenile feeding level are doing most of the work, and varying the large fish crash magnitude just changes how quickly the large fish return to a non-rate-limiting level, not the dynamics that actually drive the cycle.

## Extended Projection: Testing the Attractor

To properly test whether the fixed point (or more accurately, the limit cycle) is a genuine attractor, meaning perturbations decay back to it rather than growing or drifting, the projection time was substantially extended. Running the simulation for much longer after the perturbation gives the system enough time to either converge back to the cycle, settle to a new state, or diverge:

```{r}
#| eval: false
sim_300 <- project(sim, t_max = 200, dt = 0.1, t_save = 0.2,
                   progress_bar = FALSE, method = "predictor-corrector")

sim_600 <- project(sim_300, t_max = 300, dt = 0.1, t_save = 0.2,
                   progress_bar = FALSE, method = "predictor-corrector")
```

It converged. The trajectory returns cleanly to the same limit cycle after the perturbation, confirming that the oscillation is a stable attractor, not a transient or a neutrally stable orbit. The longer run also makes it easier to measure the period accurately and to check whether there is any amplitude modulation or beating that might indicate more than one frequency is present in the signal which, given the asymmetric shape of the feeding level animation, is a real possibility.

------------------------------------------------------------------------

## What's Next

- **FFT on the projection**: The feeding level animation and the biomass time series both show a signal that is more complicated than a simple sine wave. The asymmetric shape, long plateau, sharp crash, fast recovery, suggests the oscillation may be composed of multiple frequencies rather than a single dominant mode. A fast Fourier transform on the settled biomass (or feeding level) time series will decompose the signal and reveal whether there is one period, or several harmonics, or two incommensurate frequencies beating against each other. The answer matters: a single clean frequency implies one mechanism, while multiple frequencies would suggest the fish dynamics and the resource dynamics are on slightly different timescales that interfere.

Today was slower than expected in terms of code produced, but there were many interesting results. The feeding level animation added a layer of complexity to that picture that the FFT will hopefully help untangle.
