---
title: "Day 26: Closing Day 25's Loose Ends, and a Delay That Actually Causes a Hopf Bifurcation"
author: "Samia"
date: "2026-07-13"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, juvenile-pileup, alpha, time-delay, hopf-bifurcation, predator-prey, blog]
description: "Three loose ends from Day 25 closed: a refined alpha sweep pins the juvenile-pileup transition to a narrow (0.105, 0.178) window, the gamma = 305.8 dip turns out to flip between present and absent as t_run grows, and the tau = 40/80 amplitude climb is confirmed to decay to machine-precision zero. Separately, moving the delay onto the prey's own self-limitation (rather than the predator's conversion) produces a genuine Hopf bifurcation -- confirmed analytically and by simulation, though its amplitude drives prey biomass unphysically negative well past onset. A third variant, delaying only the predator's intake term, derives a real threshold on the prey's renewal rate but doesn't show the predicted oscillation within the delays tested."
---

## Where Day 25 Left Us

Day 25 closed the alpha/gamma/kappa juvenile-pileup sweep and ruled out a Hopf bifurcation for the time-delay predator-prey model analytically, but left three threads open: pin down alpha's transition window more precisely, check whether the gamma = 305.8 dip was real or an under-settled transient, and confirm the tau = 40/80 amplitude climb actually decays to zero given a longer `t_max` rather than plateauing. All three are closed today, in `26_experiments.R`. Along the way, a direct question came up about the delay model itself — if a time delay usually destabilises systems, why did Day 25 find it couldn't here — which led to re-deriving the Hopf condition for two different places the delay could sit, one of which does produce one.

------------------------------------------------------------------------

## Setup

```{r libraries}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse)
library(glue)
library(ggplot2)
library(colorRamps)
library(future)
library(scales)
library(deSolve)
```

Same anchovy builder and the same `catchable_fraction_at()`/`run_param_sweep()` pair from Day 25, carried over unchanged so today's refinements are directly comparable to the sweeps that found the transition and the dip in the first place:

```{r builder}
#| message: false
#| warning: false
#| output: false
make_second_order_params_kr <- function(lambda = 2.05, resource_decrease = 0.001,
                                        capacity_mult = 1, second_order = TRUE,
                                        ext_diff = 0.00, alpha = 0.1, gamma = 750,
                                        ks = 0, kappa_override = NULL) {
  a0    <- 100
  kappa <- if (is.null(kappa_override)) a0 * exp(-6.9 * (lambda - 1)) else kappa_override
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  params <- newSingleSpeciesParams(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = alpha, gamma = gamma, ks = ks
  )

  given_species_params(params)$D_ext <- ext_diff
  params <- setBevertonHolt(params)

  default_capacity <- getResourceCapacity(params)
  r  <- getResourceRate(params) * resource_decrease
  cc <- default_capacity * capacity_mult

  params <- setResource(params, resource_rate = r, resource_capacity = cc,
                        resource_dynamics = "resource_semichemostat",
                        balance = FALSE)

  if (second_order) {
    second_order_w(params) <- c(flux = "centred", bin_average = TRUE)
  }

  params
}

catchable_fraction_at <- function(params, t_max = 600) {
  sim <- project(params, t_max = t_max, dt = 0.1, t_save = 0.5,
                 progress_bar = FALSE, effort = 0, method = "tr_bdf2")
  last  <- dim(sim@n)[1]
  w_mat <- params@species_params$w_mat[1]
  w     <- sim@params@w
  dw    <- sim@params@dw

  bm_density   <- sim@n[last, 1, ] * w * dw
  total_bm     <- sum(bm_density)
  catchable_bm <- sum(bm_density[w >= w_mat])

  data.frame(catchable_fraction = catchable_bm / total_bm, error = NA_character_)
}

run_param_sweep <- function(param_name, param_seq, capacity_mult = 10,
                            resource_decrease = 0.001, t_max = 600) {
  bind_rows(lapply(param_seq, function(v) {
    args <- list(capacity_mult = capacity_mult, resource_decrease = resource_decrease)
    args[[param_name]] <- v
    result <- tryCatch({
      p <- do.call(make_second_order_params_kr, args)
      catchable_fraction_at(p, t_max = t_max)
    }, error = function(e) {
      data.frame(catchable_fraction = NA_real_, error = conditionMessage(e))
    })
    if (!is.na(result$error)) {
      warning(sprintf("%s = %.4g: %s", param_name, v, result$error))
    }
    data.frame(param = param_name, value = v,
               catchable_fraction = result$catchable_fraction,
               error = result$error)
  }))
}
```

------------------------------------------------------------------------

## Pinning Down Alpha's Transition Window

Day 25's 15-point alpha sweep found the catchable fraction jumps roughly 8 orders of magnitude somewhere between `alpha = 0.057` and `alpha = 0.28` — a genuine viability threshold, but the grid was never dense enough to say exactly where the steep part starts and ends. Same move as Day 24's `capacity_mult` refinement: 25 points, log-spaced, zoomed into `[0.05, 0.3]`, at the same `t_run = 600` so the two sweeps stay directly comparable.

```{r alpha-refine}
#| message: false
#| warning: false
alpha_seq_refine <- exp(seq(log(0.05), log(0.3), length.out = 25))
alpha_refine_df  <- run_param_sweep("alpha", alpha_seq_refine)
```

Unlike `capacity_mult`'s collapse bifurcation, there's no natural alive/dead split here — `catchable_fraction` rises smoothly (if steeply) from near-zero to a real plateau, not a bistable jump at one point. So the "transition window" is bracketed with two threshold crossings instead of one: where the climb clearly starts (still under 0.1%) and where it's clearly finished (already over 10%):

```{r find-crossing}
#| message: false
#| warning: false
#| output: false
find_crossing <- function(df, threshold, value_col = "value",
                          metric_col = "catchable_fraction") {
  df    <- df %>% arrange(.data[[value_col]])
  below <- df %>% filter(.data[[metric_col]] < threshold)  %>% pull(.data[[value_col]])
  above <- df %>% filter(.data[[metric_col]] >= threshold) %>% pull(.data[[value_col]])
  data.frame(
    threshold   = threshold,
    below_last  = if (length(below) > 0) max(below) else NA_real_,
    above_first = if (length(above) > 0) min(above) else NA_real_
  )
}

LOW_THRESHOLD  <- 1e-3
HIGH_THRESHOLD <- 0.10

alpha_crossings <- bind_rows(
  find_crossing(alpha_refine_df %>% filter(is.na(error)), LOW_THRESHOLD)  %>% mutate(edge = "climb starts"),
  find_crossing(alpha_refine_df %>% filter(is.na(error)), HIGH_THRESHOLD) %>% mutate(edge = "climb ends")
)
```

```{r alpha-plot}
#| message: false
#| warning: false
ggplot(alpha_refine_df %>% filter(is.na(error)), aes(x = value, y = catchable_fraction)) +
  geom_line(color = "#4C72B0", linewidth = 1) +
  geom_point(color = "#4C72B0", size = 1.5) +
  geom_hline(yintercept = c(LOW_THRESHOLD, HIGH_THRESHOLD), linetype = "dotted", color = "grey40") +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "grey40") +
  scale_y_log10(labels = scales::label_scientific()) +
  scale_x_log10() +
  labs(x = "alpha (assimilation efficiency)", y = "Catchable fraction (biomass at or above w_mat)",
       title = "Refined alpha sweep: pinning down the transition window",
       subtitle = "25 points over [0.05, 0.3] -- dotted = threshold crossings; dashed = anchovy's own alpha = 0.1") +
  theme_minimal()
```

| edge | bracket |
|---|---|
| Climb starts (crosses 0.1%) | `alpha ∈ (0.1055, 0.1137]` |
| Climb ends (crosses 10%) | `alpha ∈ (0.1651, 0.1779]` |

The whole steep part of the climb is compressed into a genuinely narrow window, `alpha ≈ 0.105–0.178` — much tighter than Day 25's coarse "somewhere between 0.057 and 0.28." The anchovy's own `alpha = 0.1` sits at `0.0237%` catchable fraction, just below the 0.1% "climb starts" threshold — confirming Day 25's "sits right at the foot of the climb" with an actual number behind it rather than an eyeball read.

------------------------------------------------------------------------

## Is the Gamma = 305.8 Dip Real, or an Under-Settled Transient?

Day 25's 15-point gamma sweep found one point that breaks monotonicity: `gamma = 305.8` read `9.48e-5`, dipping below both its neighbours (`231 → 2.45e-4`, `404 → 4.03e-4`) at `t_run = 600`. Flagged at the time as possibly critical slowing down — the same explanation Day 24 used for its own default-species Forward-branch spike. This reruns exactly that point and its immediate neighbours (located dynamically off the original 15-point grid, not hardcoded) at increasingly long `t_run`.

```{r gamma-dip-setup}
#| message: false
#| warning: false
#| output: false
gamma_full_seq   <- exp(seq(log(100), log(5000), length.out = 15))
dip_idx          <- which.min(abs(gamma_full_seq - 305.8))
neighbour_idx    <- unique(pmax(1, pmin(length(gamma_full_seq), c(dip_idx - 1, dip_idx, dip_idx + 1))))
gamma_check_vals <- gamma_full_seq[neighbour_idx]

t_run_check_seq <- c(600, 1200, 2400, 4800)

gamma_dip_check_df <- bind_rows(lapply(t_run_check_seq, function(tr) {
  bind_rows(lapply(gamma_check_vals, function(g) {
    p   <- make_second_order_params_kr(gamma = g, capacity_mult = 10, resource_decrease = 0.001)
    res <- catchable_fraction_at(p, t_max = tr)
    data.frame(gamma = g, t_run = tr, catchable_fraction = res$catchable_fraction)
  }))
}))

dip_persistence <- gamma_dip_check_df %>%
  group_by(t_run) %>%
  arrange(gamma, .by_group = TRUE) %>%
  summarise(
    dip_still_present = catchable_fraction[2] < catchable_fraction[1] &
                        catchable_fraction[2] < catchable_fraction[3],
    .groups = "drop"
  )
```

```{r gamma-dip-plot}
#| message: false
#| warning: false
ggplot(gamma_dip_check_df, aes(x = t_run, y = catchable_fraction, color = factor(round(gamma, 1)))) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_x_log10() +
  scale_y_log10(labels = scales::label_scientific()) +
  labs(x = "t_run", y = "Catchable fraction",
       title = "Is the gamma = 305.8 dip real or an under-settled transient?",
       subtitle = "Same point and its two neighbours, rerun at increasing t_run", color = "gamma") +
  theme_minimal()
```

| gamma | t_run=600 | t_run=1200 | t_run=2400 | t_run=4800 |
|---|---|---|---|---|
| 231.2 | 2.45e-4 | 2.19e-4 | 2.32e-4 | 2.86e-4 |
| 305.8 | 9.48e-5 | 1.45e-4 | **3.03e-4** | 9.73e-5 |
| 404.4 | 4.03e-4 | 3.39e-4 | 1.51e-4 | 1.50e-4 |

| t_run | dip still present? |
|---|---|
| 600 | TRUE |
| 1200 | TRUE |
| 2400 | **FALSE** |
| 4800 | TRUE |

This isn't "the dip shrinks and disappears" (critical slowing down) or "the dip holds steady" (a real feature) — it flips non-monotonically, and at `t_run = 2400` the ordering of all three points scrambles, not just the middle one (`305.8` becomes the *highest* of the three, and `404.4` drops below `231.2`). That's not what a converging transient looks like.

The likely explanation: `catchable_fraction_at()` doesn't average over a settled window — it reads a single snapshot at the last saved timestep of `t_max`. If the system at this operating point (`capacity_mult = 10`, `resource_decrease = 0.001` — the fixed baseline every juvenile-pileup sweep since Day 24 has used) is still gently oscillating rather than sitting at a true fixed point, then varying `t_max` just samples different *phases* of that oscillation, not different amounts of settling time. That's consistent with Day 22/23's own phase-diagram work: `capacity_mult = 10` sits just past where Day 23's refined `[3, 10]` grid found "a complete reversal" from fixed-point to oscillating. The noise floor this reveals (values bouncing 2–3x between reruns) is negligible next to alpha's 27.5-order-of-magnitude range, but it's comparable to or larger than gamma's and kappa's entire reported ranges — so those two "close to flat" / "modest effect" conclusions from Day 25 may be sampling noise from an unsettled oscillation rather than a real parameter effect. Directly checking whether this operating point is actually oscillating (a real time series, not single-snapshot reads) is the natural follow-up.

------------------------------------------------------------------------

## Confirming the Amplitude Decay for the Conversion-Delay Model

Day 25's time-delay predator-prey model (delay on the predator's conversion of intake into biomass) was proven analytically to never support a Hopf bifurcation, for any delay or parameters — but `rel_amplitude_P` was still climbing with `tau` across the tested range (up to 0.25 at `tau = 40`), which the algebra says must eventually decay rather than plateau. Reproducing Day 25's model exactly first, to confirm the base case, then extending `t_max` well past the original 300:

```{r delay-reproduce}
#| message: false
#| warning: false
#| output: false
predator_prey_params_delay <- list(D = 1, a = 1, e = 0.5, m = 0.3, K = 1)

check_hopf_feasibility <- function(parms = predator_prey_params_delay) {
  D <- parms$D; a <- parms$a; e <- parms$e; m <- parms$m; K <- parms$K
  N_star <- m / (e * a)
  P_star <- D * (K - N_star) / (a * N_star)
  alpha  <- D + a * P_star
  beta   <- a * N_star
  gamma  <- e * a * P_star
  data.frame(N_star = N_star, P_star = P_star, alpha = alpha, beta = beta,
             gamma = gamma, beta_gamma = beta * gamma, two_m_alpha = 2 * m * alpha,
             hopf_possible = (beta * gamma) > (2 * m * alpha))
}
hopf_check <- check_hopf_feasibility()

predator_prey_rhs_delay <- function(t, state, parms) {
  N <- max(state[1], 0); P <- max(state[2], 0)
  D <- parms[["D"]]; a <- parms[["a"]]; e <- parms[["e"]]
  m <- parms[["m"]]; K <- parms[["K"]]; tau <- parms[["tau"]]
  N0 <- parms[["N0"]]; P0 <- parms[["P0"]]
  if (t <= tau) { lag_N <- N0; lag_P <- P0
  } else { lag <- lagvalue(t - tau); lag_N <- max(lag[1], 0); lag_P <- max(lag[2], 0) }
  dN <- D * (K - N) - a * N * P
  dP <- e * a * lag_N * lag_P - m * P
  list(c(dN, dP))
}

run_predator_prey_delay_long <- function(tau, t_max_long, state0 = c(N = 1, P = 0.1),
                                         parms = predator_prey_params_delay) {
  parms$tau <- tau; parms$N0 <- state0[["N"]]; parms$P0 <- state0[["P"]]
  as.data.frame(dede(y = state0, times = seq(0, t_max_long, by = 0.1),
                     func = predator_prey_rhs_delay, parms = parms))
}

amplitude_at_cutoff <- function(ts_df, t_max_effective) {
  late <- ts_df[ts_df$time > t_max_effective * 0.6 & ts_df$time <= t_max_effective, ]
  data.frame(t_max = t_max_effective, rel_amplitude_P = (max(late$P) - min(late$P)) / mean(late$P))
}

tau_check_seq   <- c(40, 80)
t_max_check_seq <- c(300, 600, 1200, 2400, 4800, 9600)

ts_long_by_tau <- setNames(
  lapply(tau_check_seq, function(tau) run_predator_prey_delay_long(tau, max(t_max_check_seq))),
  tau_check_seq
)

decay_check_df <- bind_rows(lapply(tau_check_seq, function(tau) {
  ts_long <- ts_long_by_tau[[as.character(tau)]]
  bind_rows(lapply(t_max_check_seq, function(tm) amplitude_at_cutoff(ts_long, tm) %>% mutate(tau = tau)))
}))
```

```{r decay-plot}
#| message: false
#| warning: false
ggplot(decay_check_df, aes(x = t_max, y = rel_amplitude_P, color = factor(tau))) +
  geom_line(linewidth = 1) + geom_point(size = 1.5) +
  scale_x_log10() + scale_y_log10(labels = scales::label_scientific()) +
  labs(x = "t_max", y = "Relative amplitude of P, late window",
       title = "Does rel_amplitude_P decay to ~0 given enough time, or plateau?",
       subtitle = "tau = 40 and 80, re-windowed at longer cutoffs -- analytic prediction: decays to 0",
       color = "tau") +
  theme_minimal()
```

| t_max | tau=40 | tau=80 |
|---|---|---|
| 300 | 0.2495 | 0.2513 |
| 600 | 0.0484 | 0.2448 |
| 1200 | 8.00e-4 | 0.0463 |
| 2400 | 1.51e-7 | 5.18e-4 |
| 4800 | 4.16e-15 | 8.34e-8 |
| 9600 | 1.67e-16 | 1.67e-15 |

Both `tau` values decay across 15 orders of magnitude to machine-precision zero — not a plateau. A separate endpoint check (distance from the trajectory's final `N, P` to the analytic equilibrium) confirms it directly: both `tau = 40` and `tau = 80` land within `3.7e-15` of `(N^*, P^*)` by `t_max = 9600`. That closes Day 25's third loose end: the climbing amplitude sweep was measuring an unfinished monotonic relaxation, not a real or growing oscillation, exactly as the algebra predicted.

------------------------------------------------------------------------

## Why Couldn't Delay Cause an Instability Here?

Day 25's model delays the predator's *conversion* of intake into biomass, but leaves the prey's own self-limitation instantaneous:

```
dN/dt = D(K - N(t))       - aN(t)P(t)
dP/dt = eaN(t-tau)P(t-tau) - mP(t)
```

Linearising around equilibrium and substituting `n, p ~ e^(lambda t)` gives the characteristic equation `(lambda+alpha)(lambda+m-me^(-lambda tau)) + beta gamma e^(-lambda tau) = 0`, with `alpha = D + aP^*`, `beta = aN^*`, `gamma = eaP^*`. Setting `lambda = i omega` (the only way a root can cross from stable to unstable as `tau` grows) and matching `|P(i omega)| = |Q(i omega)|` reduces this to a quadratic in `u = omega^2`:

```
u^2 + alpha^2 u + beta gamma (2 m alpha - beta gamma) = 0
```

By Vieta's formulas, the sum of the two roots is `-alpha^2` (always negative), so a positive root — a genuine crossing frequency — can only exist if the *product*, `beta gamma (2 m alpha - beta gamma)`, is negative, i.e. if `beta gamma > 2 m alpha`. Substituting the actual definitions and simplifying (using the equilibrium relation `eaN^* = m`):

```
beta gamma - 2 m alpha = -a m P^* - 2 m D
```

Factoring out `-mD` (after substituting `P^* = D(K-N^*)/(aN^*)`) gives a strikingly clean closed form: `beta gamma - 2 m alpha = -mD(K+N^*)/N^*`. Every quantity in that bracket is positive, so `D` sits as a bare positive multiplier on an unconditionally negative expression — for *any* positive `D, a, e, m, K`, this can never be positive. The undelayed prey renewal term `D` is doing all the work: it's a restoring force that acts at full strength regardless of `tau`, and it's strong enough here to prevent any Hopf bifurcation, no matter the delay.

That raised an obvious question: what if the delay sat somewhere else?

------------------------------------------------------------------------

## A Model That Does Support a Hopf Bifurcation: Delay the Prey's Self-Limitation Instead

Moving the delay onto the prey's own self-limitation term instead — `dN/dt = D(K - N(t-tau)) - aN(t)P(t)`, predation and conversion now instantaneous — is the classical route to delay-induced instability, the two-species analogue of Hutchinson's single-species delayed logistic equation.

### Working Out the Maths

**Step 1 — the equilibrium is unchanged.** At a constant solution the delayed term equals its instantaneous value, so `N^* = m/(ea)`, `P^* = D(K-N^*)/(aN^*)` — identical to before.

**Step 2 — linearise.** Writing `N = N^*+n`, `P = P^*+p` and dropping second-order terms gives:

```
dn/dt = -alpha n(t) - beta p(t) - D n(t-tau)     [alpha = a P^*, NOT D+aP^* -- D now
                                                    sits on its own, as the coefficient
                                                    of the delayed term]
dp/dt = gamma n(t)                                [gamma = eaP^*, no delay at all here]
```

The key structural change from before: `D` has been pulled out of `alpha` entirely and now multiplies the delayed term directly.

**Step 3 — the characteristic equation.** Substituting the ansatz gives `lambda^2 + alpha lambda + beta gamma + D lambda e^(-lambda tau) = 0`. At `tau=0` this becomes `lambda^2 + (alpha+D)lambda + beta gamma = 0`, and `alpha+D = aP^*+D` is exactly the *old* `alpha` — matching the non-delayed result exactly, as it must.

**Step 4 — matching moduli.** Setting `lambda = i omega` and matching `|P(i omega)| = |Q(i omega)|` (same method as before) gives a quadratic in `u = omega^2`:

```
u^2 + (alpha^2 - D^2 - 2 beta gamma) u + (beta gamma)^2 = 0
```

**Step 5 — the sign test.** The product of roots here is `(beta gamma)^2` — always a perfect square, always non-negative. That rules out the "opposite signs" case from the earlier model entirely: both roots always share the same sign, and the sum decides which. Requiring both real (discriminant ≥ 0) and positive (sum > 0), and simplifying, collapses to one clean condition:

```
Hopf possible  <=>  alpha < D  <=>  a P^* < D
```

Checked against this project's own reference parameters (`D=1, a=1, e=0.5, m=0.3, K=1` → `P^*=0.667`): `a P^* = 0.667 < D = 1` — the condition **holds**, with the exact same numbers used throughout Days 25–26.

### Confirming It Numerically

```{r selflim-setup}
#| message: false
#| warning: false
#| output: false
predator_prey_rhs_selflim <- function(t, state, parms) {
  N <- max(state[1], 0); P <- max(state[2], 0)
  D <- parms[["D"]]; a <- parms[["a"]]; e <- parms[["e"]]
  m <- parms[["m"]]; K <- parms[["K"]]; tau <- parms[["tau"]]; N0 <- parms[["N0"]]
  lag_N <- if (t <= tau) N0 else max(lagvalue(t - tau, 1), 0)
  dN <- D * (K - lag_N) - a * N * P
  dP <- e * a * N * P - m * P
  list(c(dN, dP))
}

check_hopf_feasibility_selflim <- function(parms = predator_prey_params_delay) {
  D <- parms$D; a <- parms$a; e <- parms$e; m <- parms$m; K <- parms$K
  N_star  <- m / (e * a)
  P_star  <- D * (K - N_star) / (a * N_star)
  data.frame(N_star = N_star, P_star = P_star, alpha_s = a * P_star, D = D,
             hopf_possible = (a * P_star) < D)
}

run_predator_prey_selflim <- function(tau, state0 = c(N = 1, P = 0.1), t_max = 300,
                                      parms = predator_prey_params_delay) {
  parms$tau <- tau; parms$N0 <- state0[["N"]]
  out  <- as.data.frame(dede(y = state0, times = seq(0, t_max, by = 0.1),
                             func = predator_prey_rhs_selflim, parms = parms))
  late <- out[out$time > t_max * 0.6, ]
  data.frame(tau = tau, max_N = max(late$N), min_N = min(late$N),
             max_P = max(late$P), min_P = min(late$P),
             rel_amplitude_P = (max(late$P) - min(late$P)) / mean(late$P))
}

tau_seq_selflim  <- c(0, 0.5, 1, 1.5, 2, 2.2, 2.4, 2.6, 2.8, 3, 4, 5, 7, 10, 15, 20)
selflim_sweep_df <- bind_rows(lapply(tau_seq_selflim, run_predator_prey_selflim))
```

```{r selflim-plot}
#| message: false
#| warning: false
ggplot(selflim_sweep_df, aes(x = tau, y = rel_amplitude_P)) +
  geom_line(color = "#C44E52", linewidth = 1) + geom_point(color = "#C44E52", size = 1.5) +
  labs(x = "tau (delay in prey self-limitation)", y = "Relative amplitude of P (settled)",
       title = "Delaying the prey's self-limitation: a genuine Hopf bifurcation appears",
       subtitle = "Same D=1,a=1,e=0.5,m=0.3,K=1 -- analytic condition a*P*=0.667 < D=1 predicts yes") +
  theme_minimal()
```

| tau | rel_amplitude_P | min_N |
|---|---|---|
| 2.2 | 0.0204 | 0.579 |
| **2.4** | **0.661** | **-0.021** |
| 2.6 | 0.827 | -0.225 |
| 3 | 1.047 | -0.548 |
| 5 | 2.029 | -2.346 |
| 10 | 3.059 | -6.956 |
| 20 | 4.214 | -15.532 |

The onset is sharp — between `tau = 2.2` (amplitude still `0.02`, essentially flat) and `tau = 2.4` (amplitude jumps to `0.66`) — landing almost exactly on the `tau ≈ 2.4` estimate worked out by hand from the phase-matching condition beforehand. A persistence check at `tau = 10` across `t_max = 300, 1500, 3000` gives `rel_amplitude_P = 3.06, 2.95, 2.92` — steady, not decaying, confirming this is a genuine sustained limit cycle rather than another unsettled transient.

**But there's a real problem with it.** Look at the `min_N` column: `N` (prey biomass) goes negative right at onset and gets dramatically more negative as `tau` grows — down to `-15.5` by `tau = 20`. This toy model has no mechanism (no saturating functional response, no higher-order term) to cap the amplitude of the resulting oscillation, so once you're more than a little past the bifurcation point, the "population" swings through biologically nonsensical negative values. The Hopf bifurcation itself is real and confirmed both analytically and numerically — but this specific model can't be trusted as a faithful biological picture much beyond its onset.

------------------------------------------------------------------------

## A Third Variant: Delay Only the Predator's Intake, Not Its Own Growth

A natural intermediate case: what if only `P(t-tau)` is delayed inside the predator's growth term, while `N(t)` there stays instantaneous?

```
dN/dt = D(K - N(t))       - aN(t)P(t)
dP/dt = eaN(t)P(t-tau)     - mP(t)
```

### The Maths

Linearising gives `dp/dt = gamma n(t) + m p(t-tau) - m p(t)` — the `n` term loses its delay, but the `m p(t-tau)` term survives (it comes from `N^*` — a constant — multiplying the still-delayed `P(t-tau)`). The characteristic equation becomes `(lambda+alpha)(lambda+m-me^(-lambda tau)) + beta gamma = 0`, with `alpha = D+aP^*` as in the original model, but now `beta gamma` carries no `e^(-lambda tau)` factor at all.

Matching moduli gives `u^2 + (alpha^2 - 2 beta gamma) u + beta gamma(2 alpha m + beta gamma) = 0`. The product of roots, `beta gamma(2 alpha m + beta gamma)`, is always positive (every factor is positive) — same situation as the self-limitation model, so both roots share a sign and the sum decides which:

```
Hopf possible  <=>  2 beta gamma > alpha^2
```

Writing `P^* = D*C` (`C = (K-N^*)/(aN^*)`, independent of `D`) and substituting through reduces this to a genuine **threshold on D**, unlike the absolute impossibility of the original model and unlike the self-limitation model's condition (which held outright at the reference parameters):

```
D < 2 a m C / (1 + a C)^2
```

At the reference parameters (`N^*=0.6, C=0.667`), that threshold works out to `~0.144`. `D=1` is far above it, so no Hopf is predicted at the reference `D` — tested below alongside `D=0.1`, comfortably under the threshold.

### Confirming It Numerically

```{r pdelay-setup}
#| message: false
#| warning: false
#| output: false
predator_prey_rhs_pdelay <- function(t, state, parms) {
  N <- max(state[1], 0); P <- max(state[2], 0)
  D <- parms[["D"]]; a <- parms[["a"]]; e <- parms[["e"]]
  m <- parms[["m"]]; K <- parms[["K"]]; tau <- parms[["tau"]]; P0 <- parms[["P0"]]
  lag_P <- if (t <= tau) P0 else max(lagvalue(t - tau, 2), 0)
  dN <- D * (K - N) - a * N * P
  dP <- e * a * N * lag_P - m * P
  list(c(dN, dP))
}

run_predator_prey_pdelay <- function(tau, state0 = c(N = 1, P = 0.1), t_max = 300,
                                     parms = predator_prey_params_delay) {
  parms$tau <- tau; parms$P0 <- state0[["P"]]
  out  <- as.data.frame(dede(y = state0, times = seq(0, t_max, by = 0.1),
                             func = predator_prey_rhs_pdelay, parms = parms))
  late <- out[out$time > t_max * 0.6, ]
  data.frame(tau = tau, max_N = max(late$N), min_N = min(late$N),
             max_P = max(late$P), min_P = min(late$P),
             rel_amplitude_P = (max(late$P) - min(late$P)) / mean(late$P))
}

params_small_D  <- modifyList(predator_prey_params_delay, list(D = 0.1))
tau_seq_pdelay  <- c(0, 0.5, 1, 2, 3, 5, 7, 10, 15, 20)

pdelay_sweep_refD   <- bind_rows(lapply(tau_seq_pdelay, run_predator_prey_pdelay)) %>%
  mutate(D_label = "D = 1 (above threshold)")
pdelay_sweep_smallD <- bind_rows(lapply(tau_seq_pdelay, run_predator_prey_pdelay,
                                        parms = params_small_D)) %>%
  mutate(D_label = "D = 0.1 (below threshold)")
pdelay_sweep_df <- bind_rows(pdelay_sweep_refD, pdelay_sweep_smallD)
```

```{r pdelay-plot}
#| message: false
#| warning: false
ggplot(pdelay_sweep_df, aes(x = tau, y = rel_amplitude_P, color = D_label)) +
  geom_line(linewidth = 1) + geom_point(size = 1.5) +
  labs(x = "tau (delay in P(t-tau) term only, N instantaneous)", y = "Relative amplitude of P",
       title = "Delaying only P inside the intake term: a threshold on D, not an absolute block",
       subtitle = "Derived threshold D < 0.144 -- tested on both sides", color = NULL) +
  theme_minimal()
```

At `D=1` (above threshold), `rel_amplitude_P` climbs smoothly from `1.8e-10` to `0.145` by `tau=20` — the same shape as the original conversion-delay model's non-Hopf climb, as expected. At `D=0.1` (below threshold, Hopf predicted), the climb is smaller and equally gradual, reaching only `0.033` by `tau=20`, with no sharp transition visible anywhere in the tested range.

A persistence check at `D=0.1, tau=10` settles it: `rel_amplitude_P` at `t_max = 300, 1500, 3000` is `0.00164 → 9.18e-14 → 1.04e-15` — collapsing toward machine-precision zero. That's the exact "critical slowing down, not real oscillation" signature from earlier, not a sustained cycle. So even though `D=0.1` genuinely satisfies the derived existence condition, `tau=10` isn't past the real onset for this variant — the true critical `tau` must sit somewhere beyond 20, not yet located. This is a real, if unglamorous, result: the analytic threshold on `D` is correctly derived and internally consistent, but pinning down *where* (in `tau`) the bifurcation actually occurs for a given `D` needs a wider sweep than was run today.

------------------------------------------------------------------------

## What This All Adds Up To

Three loose ends from Day 25 closed cleanly: alpha's transition window is now a precise `(0.105, 0.178)` bracket, the tau = 40/80 amplitude climb is confirmed to decay to machine-precision zero rather than plateau, and the gamma = 305.8 dip turned out to be more interesting than either "real" or "transient" — its non-monotonic flipping points at an unsettled oscillation at the sweep's own operating point, which puts a question mark over gamma and kappa's "modest effect" conclusions from Day 25 that alpha's huge range was never at risk from.

The deeper result is about *where* a delay sits, not whether delay can destabilise a system in general. Delaying the predator's conversion (Day 25's model) can never produce a Hopf bifurcation here, for any parameters — proven by factoring the destabilising quantity into `-mD(K+N^*)/N^*`, a bare positive `D` multiplying an unconditionally negative bracket. Delaying the prey's own self-limitation instead does produce one, confirmed both by rederiving the condition (`aP^* < D`) and by simulation landing the onset almost exactly where the hand-derived estimate predicted — though the resulting limit cycle's amplitude grows unboundedly past onset, driving prey biomass negative, a real limitation of this toy model rather than of the underlying mechanism. A third, intermediate variant (delay only the predator's intake, not its own growth) derives a genuine threshold on `D` rather than an absolute block, but the specific `tau` tested wasn't far enough to actually see it.

------------------------------------------------------------------------

## What's Next

- **Find the actual critical tau for the P-only-delay variant at D=0.1.** Extend `tau_seq_pdelay` well past 20 (e.g. up to 40–60) and rerun the persistence check at each candidate onset to locate where `rel_amplitude_P` stops decaying and starts holding steady.
- **Characterise how far past onset the self-limitation model's amplitude stays biologically sensible.** `min_N` is already negative at `tau=2.4` (barely past onset) and reaches `-15.5` by `tau=20` — worth finding the `tau` range where `N` stays non-negative, and considering whether a saturating term (a proper functional response, rather than mass-action) would cap the amplitude the way it should.
- **Check directly whether `capacity_mult=10, resource_decrease=0.001` is actually oscillating**, rather than reading single-timestep snapshots at different `t_max` — a real time series at this exact operating point would settle whether the gamma/kappa sweep's noise floor is genuine unsettled oscillation or something else.
- **Rerun gamma and kappa's isolation sweeps with a windowed-average read** (the same "late window" convention used throughout the delay-model work) instead of a single final-timestep snapshot, to check whether their "modest"/"flat" Day 25 conclusions survive averaging out whatever phase-sampling noise today's gamma recheck exposed.
