---
title: "Day 19: balance = FALSE Doesn't Kill the Oscillation, It Starves the Population"
author: "Samia"
date: "2026-07-02"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, limit-cycle, resource-balance, bifurcation, yield, blog]
description: "Re-running Day 18's resource_decrease sweep with balance = FALSE first looks like it kills the oscillation, but a controlled A/B shows it's actually chronic starvation, not stabilisation. Fixing the resource baseline at its best-oscillating values and varying the feeding kernel instead shows kernel shape sets oscillation amplitude directly, and explains an 8x yield gap between two kernels: most of the higher-amplitude kernel's extra biomass is juveniles below the gear's catchable-size cutoff."
---

## Where Day 18 Left Us

Day 18 confirmed one thing outright: the limit cycle is a stable global attractor (trajectories from perturbations spanning three orders of magnitude all converge to the same orbit). The bifurcation's character was left open rather than settled: amplitude grows continuously from zero on the forward sweep (a supercritical signature), but the backward sweep sustains oscillation to a noticeably higher `resource_decrease` than the forward onset — a gap too large to write off as a sweep-start artefact alone, so a purely supercritical Hopf bifurcation couldn't yet be confirmed. Its "What's Next" list left several threads open: try a different perturbation type, run with `balance = FALSE` in the resource-setting step, plot maximum and minimum separately instead of max − min, and resolve whether that forward/backward gap is real hysteresis or a run-length artefact.

Today tackles the `balance = FALSE` thread, and it took a couple of false starts to understand what's actually going on.

```{r}
#| message: false
#| warning: false
#| output: false
library(mizer)
library(patchwork)
library(reshape2)
library(plotly)
library(dplyr)
library(mizerExperimental)
library(tidyverse)
library(glue)
library(ggplot2)
library(colorRamps)
library(future.apply)
```

------------------------------------------------------------------------

## Setup

`balance` is now an explicit argument on `make_params`, defaulting to `FALSE`. `r_d` is the same log-spaced grid from 0.001 to 0.3 used in Day 18's hysteresis sweep — it brackets the bifurcation point found there (`rd ≈ 0.008`), so every value tested produced clear oscillation under the default `balance = TRUE` setting.

```{r}
#| message: false
#| warning: false
#| output: false
make_params <- function(lambda = 2.05, resource_decrease = 0.001,
                        beta = NULL, sigma = NULL, balance = FALSE) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  args <- list(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )
  if (!is.null(beta))  args$beta  <- beta
  if (!is.null(sigma)) args$sigma <- sigma

  params <- do.call(newSingleSpeciesParams, args)
  r      <- getResourceRate(params) * resource_decrease
  setResource(params, resource_rate = r,
             resource_dynamics = "resource_semichemostat", balance = balance)
}

make_limit_cycle_sim <- function(params, t_total = 600, effort = 0, perturbation = 1e3) {
  params@initial_n_pp[] <- params@cc_pp * 0.1

  sim_init <- project(params, t_max = 10, dt = 0.1, t_save = 0.2,
                      progress_bar = FALSE, effort = 0,
                      method = "predictor-corrector")
  idx  <- params@w >= 10 & params@w <= 100
  last <- dim(sim_init@n)[1]
  sim_init@n[last, , idx] <- sim_init@n[last, , idx] / perturbation

  project(sim_init, t_max = t_total - 10, dt = 0.1, t_save = 0.2,
          progress_bar = FALSE, effort = effort,
          method = "tr_bdf2")
}

test_effort <- function(r_d, lambda = 2.05, t_total = 600, balance = FALSE) {
  p   <- make_params(lambda = lambda, resource_decrease = r_d, balance = balance)
  sim <- make_limit_cycle_sim(p, t_total = t_total)
  bm  <- getBiomass(sim)[, "Anchovy"]
  t   <- as.numeric(names(bm))
  list(sim = sim, bm = bm, t = t)
}

r_d <- exp(seq(log(0.001), log(0.3), length.out = 10))
```

------------------------------------------------------------------------

## Experiment: Does `balance = FALSE` Kill the Oscillation?

The first pass just re-runs Day 18's sweep with the new default (`balance = FALSE`):

```{r}
#| message: false
#| warning: false
#| output: false
plan(multisession)
results_fishing <- future_lapply(r_d, test_effort, future.seed = TRUE)
```

```{r}
#| message: false
#| warning: false
plot(NULL, xlim = c(550, 600),
     ylim = range(sapply(results_fishing, function(r) range(r$bm[r$t > 550]))),
     xlab = "Time (years)", ylab = "Biomass", main = "Biomass trajectories (t > 550), balance = FALSE")
cols <- colorRampPalette(c("blue", "red"))(length(r_d))
for (i in seq_along(results_fishing)) {
  r    <- results_fishing[[i]]
  mask <- r$t > 550
  lines(r$t[mask], r$bm[mask], col = cols[i])
}
legend("topright", legend = paste0("r_d = ", r_d), col = cols, lty = 1, cex = 0.8)
```

Every trajectory is flat by $t = 550$, across the entire range that produced clear oscillation under `balance = TRUE` on Day 18. Read at face value, this says `balance = FALSE` removes the oscillatory mechanism. That reading turns out to be wrong.

A controlled A/B settles it: the same grid, run with `balance = TRUE` and `balance = FALSE`, holding every other setting (solvers, perturbation, `t_total`) identical.

```{r}
#| message: false
#| warning: false
#| output: false
plan(multisession)
results_balance_true  <- future_lapply(r_d, test_effort, balance = TRUE,  future.seed = TRUE)
results_balance_false <- future_lapply(r_d, test_effort, balance = FALSE, future.seed = TRUE)

amp_by_balance <- data.frame(
  r_d = r_d,
  amp_balance_true  = sapply(results_balance_true,  function(r) {
    late <- r$bm[r$t > 550]; max(late) - min(late)
  }),
  amp_balance_false = sapply(results_balance_false, function(r) {
    late <- r$bm[r$t > 550]; max(late) - min(late)
  })
)
```

```{r}
#| message: false
#| warning: false
knitr::kable(amp_by_balance, digits = 20)
```

`amp_balance_true` reproduces Day 18's bifurcation exactly: large amplitude (0.05–0.16) below `r_d ≈ 0.008–0.02`, then a hard drop to `1e-16`-scale floating-point noise above it, which shows that it is stable fixed point. `amp_balance_false` never gets near that amplitude, but it doesn't look like flat noise either: it climbs smoothly across roughly sixteen orders of magnitude, from `4.5e-20` at `r_d = 0.001` up to `1e-4` at `r_d = 0.3`. Floating-point noise around a fixed point doesn't behave like that but a quantity that scales smoothly with its own magnitude does, which is what it looks like if the biomass itself is climbing steadily across the grid rather than sitting at a fixed, non-oscillating level.

Checking final biomass directly, not just amplitude, confirms it:

```{r}
#| message: false
#| warning: false
bm_final_balance <- data.frame(
  r_d            = r_d,
  bm_final_true  = sapply(results_balance_true,  function(r) tail(r$bm, 1)),
  bm_final_false = sapply(results_balance_false, function(r) tail(r$bm, 1))
)
knitr::kable(bm_final_balance, digits = 20)
```

At `r_d = 0.001`, `bm_final_false` is `1.6e-17` , the population is extinct , against `bm_final_true`'s real biomass of `0.025`. It isn't just the smallest `r_d` either: even at `r_d = 0.3`, the top of the entire tested range, `bm_final_false` (`1.1e-3`) is still roughly 76 times smaller than `bm_final_true`'s stable-fixed-point plateau (`0.087`). The population under `balance = FALSE` never reaches a biomass comparable to the `balance = TRUE` case anywhere in this grid, let alone the level needed to destabilise into a limit cycle.

### What This Shows

`balance = FALSE` doesn't remove the oscillatory mechanism. It chronically underfeeds the population across the whole `resource_decrease` range, and drives it toward extinction at the low end. The two settings were never testing comparable resource availability at the "same" `resource_decrease` value, `resource_decrease = 0.3` under `balance = FALSE` leaves the population worse off than `resource_decrease = 0.001` does under `balance = TRUE`.

A tempting shortcut here would have been to check `resource_capacity` (`cc_pp`) directly on a freshly built `params` object: it comes back identical between `balance = TRUE` and `balance = FALSE`, which looks like it rules capacity out entirely. It doesn't so that check runs before any consumption dynamics exist for `balance` to compensate for, so its real effect only shows up once the model actually runs, as the biomass comparison above confirms.

------------------------------------------------------------------------

## Experiment: Feeding-Kernel Shape, Fishing Effort, and the Yield Gap

A first attempt to find *some* destabilising effect under `balance = FALSE` scanned the feeding kernel (`beta`, `sigma`) at `resource_decrease` fixed at `0.001` and found nothing, every cell essentially zero amplitude. That's explained by the previous section: at `r_d = 0.001` under `balance = FALSE` the population is already deep in the near-extinct regime, and no feeding-kernel shape can matter when there's essentially no population left to feed.

Repeating it properly means fixing `resource_rate`/`resource_capacity` at values that are actually known to sustain oscillation, pulled directly from the strongest `balance = TRUE` run above (`r_d = r_d[1]`, amplitude `0.159`), and varying `beta`/`sigma` on top of that known-good baseline instead of a starved one.

```{r}
#| message: false
#| warning: false
#| output: false
best_params   <- make_params(resource_decrease = r_d[1], balance = TRUE)
best_rate     <- best_params@rr_pp
best_capacity <- best_params@cc_pp

# Supplying both resource_rate and resource_capacity leaves nothing for
# balance to back-solve, so balance = FALSE here just sets them directly.
make_params_fixed_resource <- function(lambda = 2.05, resource_rate, resource_capacity,
                                       beta = NULL, sigma = NULL) {
  a0    <- 100
  kappa <- a0 * exp(-6.9 * (lambda - 1))
  no_w  <- round(log(66.5 / 0.0003) / 0.1)

  args <- list(
    species_name = "Anchovy",
    w_min = 0.0003, w_max = 66.5, w_mat = 10,
    no_w = no_w, lambda = lambda, kappa = kappa,
    alpha = 0.1, gamma = 750, ks = 0
  )
  if (!is.null(beta))  args$beta  <- beta
  if (!is.null(sigma)) args$sigma <- sigma

  params <- do.call(newSingleSpeciesParams, args)
  setResource(params, resource_rate = resource_rate, resource_capacity = resource_capacity,
             resource_dynamics = "resource_semichemostat", balance = FALSE)
}

test_beta_sigma_fixed <- function(beta, sigma, resource_rate, resource_capacity,
                                  lambda = 2.05, t_total = 600) {
  p   <- make_params_fixed_resource(lambda = lambda, resource_rate = resource_rate,
                                    resource_capacity = resource_capacity,
                                    beta = beta, sigma = sigma)
  sim <- make_limit_cycle_sim(p, t_total = t_total)
  bm  <- getBiomass(sim)[, "Anchovy"]
  t   <- as.numeric(names(bm))
  list(sim = sim, bm = bm, t = t, beta = beta, sigma = sigma)
}

# newSingleSpeciesParams() defaults to beta = 100, sigma = 1.3, so the grid
# is centred on those defaults with room to move in both directions.
beta_seq_fixed  <- c(50, 75, 100, 150, 200, 300)
sigma_seq_fixed <- c(0.8, 1.0, 1.3, 1.6, 2.0)
grid_fixed      <- expand.grid(beta = beta_seq_fixed, sigma = sigma_seq_fixed)

plan(multisession)
results_beta_sigma_fixed <- future_lapply(seq_len(nrow(grid_fixed)), function(i) {
  test_beta_sigma_fixed(beta = grid_fixed$beta[i], sigma = grid_fixed$sigma[i],
                        resource_rate = best_rate, resource_capacity = best_capacity)
}, future.seed = TRUE)

amp_df_fixed <- data.frame(
  beta  = grid_fixed$beta,
  sigma = grid_fixed$sigma,
  amp   = sapply(results_beta_sigma_fixed, function(r) {
    late <- r$bm[r$t > 550]
    max(late) - min(late)
  })
)
```

```{r}
#| message: false
#| warning: false
ggplot(amp_df_fixed, aes(x = factor(beta), y = factor(sigma), fill = amp)) +
  geom_tile() +
  geom_point(data = subset(amp_df_fixed, beta == 100 & sigma == 1.3),
             aes(x = factor(beta), y = factor(sigma)),
             shape = 4, size = 4, color = "red") +
  scale_fill_viridis_c() +
  labs(x = "beta", y = "sigma",
       fill = "Amplitude\n(max - min, t > 550)",
       title = "Oscillation amplitude across beta/sigma grid",
       subtitle = "resource_rate/capacity fixed at the best-oscillating balance = TRUE values; red x = defaults (beta=100, sigma=1.3)") +
  theme_minimal()
```

The red-marked cell is the default kernel (`beta = 100, sigma = 1.3`) behind every oscillation characterised in Days 10–18. This heatmap has real structure: amplitude ranges smoothly from around `0.05` (dark, top-right) up to over `0.5` (bright yellow, `beta = 50, sigma = 0.8`), with the default sitting mid-range around `0.25`–`0.3`. Narrower, more concentrated feeding kernels (lower `beta`, lower `sigma`) sustain a noticeably larger-amplitude oscillation than the default; wider ones damp it down.

Checking the actual biomass trajectories behind the darkest cells (not shown in full here) rules out one obvious confound: most low-amplitude cells are near-fixed-points, not starvation. But a couple are still slowly relaxing toward their eventual level rather than truly flat by `t = 600`, and one (`beta = 150, sigma = 0.8`) develops irregular, non-periodic fluctuation on a much longer run, a possible chaos signature, parked here for a proper Poincaré-map test (see Theoretical Grounding below).

### Fishing Effort in the Oscillating Regime

The more immediately useful question is what fishing does to yield in the parts of this grid that are unambiguously oscillating: the default kernel (`beta = 100, sigma = 1.3`) and the highest-amplitude one (`beta = 50, sigma = 0.8`), both run across a range of `effort` values on the same fixed best-oscillating resource baseline. `effort = 0` (no fishing) is the baseline every fished run is compared against.

```{r}
#| message: false
#| warning: false
#| output: false
test_effort_fixed <- function(effort, beta, sigma, resource_rate, resource_capacity,
                              lambda = 2.05, t_total = 600) {
  p     <- make_params_fixed_resource(lambda = lambda, resource_rate = resource_rate,
                                      resource_capacity = resource_capacity,
                                      beta = beta, sigma = sigma)
  sim   <- make_limit_cycle_sim(p, t_total = t_total, effort = effort)
  bm    <- getBiomass(sim)[, "Anchovy"]
  yield <- getYield(sim)[, "Anchovy"]
  t     <- as.numeric(names(bm))
  list(sim = sim, bm = bm, yield = yield, t = t,
       effort = effort, beta = beta, sigma = sigma)
}

effort_seq <- seq(0, 6, length.out = 13)  # extended from 0-2: both curves were still climbing at effort = 2

fishing_kernels <- list(
  list(beta = 100, sigma = 1.3, label = "default (beta=100, sigma=1.3)"),
  list(beta = 50,  sigma = 0.8, label = "highest-amplitude (beta=50, sigma=0.8)")
)

plan(multisession)
fishing_results <- lapply(fishing_kernels, function(k) {
  runs <- future_lapply(effort_seq, function(e) {
    test_effort_fixed(effort = e, beta = k$beta, sigma = k$sigma,
                      resource_rate = best_rate, resource_capacity = best_capacity)
  }, future.seed = TRUE)
  list(kernel = k, runs = runs)
})

fishing_summary <- bind_rows(lapply(fishing_results, function(fr) {
  data.frame(
    kernel     = fr$kernel$label,
    effort     = effort_seq,
    mean_bm    = sapply(fr$runs, function(r) mean(r$bm[r$t > 550])),
    mean_yield = sapply(fr$runs, function(r) mean(r$yield[r$t > 550])),
    amp        = sapply(fr$runs, function(r) {
      late <- r$bm[r$t > 550]; max(late) - min(late)
    })
  )
}))
```

```{r}
#| message: false
#| warning: false
ggplot(fishing_summary, aes(x = effort, y = mean_yield, color = kernel)) +
  geom_line(linewidth = 1) + geom_point() +
  labs(x = "Fishing effort", y = "Mean yield (t > 550)",
       title = "Yield vs fishing effort in the oscillating regime",
       subtitle = "effort = 0 (leftmost point) is the unfished baseline") +
  theme_minimal()
```

```{r}
#| message: false
#| warning: false
baseline_bm <- fishing_summary[fishing_summary$effort == 0, c("kernel", "mean_bm")]
names(baseline_bm)[2] <- "baseline_bm"

ggplot(fishing_summary, aes(x = effort, y = mean_bm, color = kernel)) +
  geom_line(linewidth = 1) + geom_point() +
  geom_hline(data = baseline_bm, aes(yintercept = baseline_bm, color = kernel),
             linetype = "dashed") +
  labs(x = "Fishing effort", y = "Mean biomass (t > 550)",
       title = "Biomass vs fishing effort in the oscillating regime",
       subtitle = "Dashed lines = each kernel's unfished (effort = 0) baseline biomass") +
  theme_minimal()
```

```{r}
#| message: false
#| warning: false
ggplot(fishing_summary, aes(x = effort, y = amp, color = kernel)) +
  geom_line(linewidth = 1) + geom_point() +
  labs(x = "Fishing effort", y = "Oscillation amplitude (max - min, t > 550)",
       title = "Does fishing effort dampen the oscillation?") +
  theme_minimal()
```

`effort_seq` was extended from `0–2` to `0–6` after a first pass, since both yield curves were still climbing at `effort = 2`. Biomass declines smoothly for both kernels, the highest-amplitude kernel drops about 44% from its unfished baseline (`0.225 → 0.125`), the default about 28% (`0.097 → 0.07`), but neither collapses to zero even by `effort = 6`, and yield still hasn't peaked for either: the default flattens near the top (\~`1.05e-4`) but hasn't turned down, and the highest-amplitude kernel has essentially plateaued around `1.5e-5`. Since neither stock is collapsing, there's no maximum-sustainable-yield hump showing up in this range; as fishing reduces the fish population, predation pressure on the resource likely eases and lets it partially recover, buffering yield and pushing any real hump further out than `effort = 6`. The wider range also resolves an earlier ambiguity about damping: the default kernel's amplitude does decline with effort (`0.16 → ~0.08`, roughly halved), it just needed the wider window to separate the trend from noise; the highest-amplitude kernel's decline is smooth and monotonic throughout (`0.6 → 0.27`).

Most striking: **the highest-amplitude kernel has more than double the default kernel's unfished biomass (\~0.22 vs \~0.097) but produces roughly 8× less yield at every effort level tested.** More standing stock is not turning into more catch. Gear selectivity here is knife-edge at `w_mat` by default, meaning only fish above that size are catchable.So if the highest-amplitude kernel's biomass is concentrated below that cutoff, that alone explains the mismatch. This is checkable for free, reusing the `sim` objects already stored from the `effort = 0` runs above:

```{r}
#| message: false
#| warning: false
unfished_default     <- fishing_results[[1]]$runs[[which(effort_seq == 0)]]$sim
unfished_highest_amp <- fishing_results[[2]]$runs[[which(effort_seq == 0)]]$sim

last_default <- dim(unfished_default@n)[1]
last_highamp <- dim(unfished_highest_amp@n)[1]

spectrum_df <- bind_rows(
  data.frame(
    w               = unfished_default@params@w,
    biomass_density = unfished_default@n[last_default, 1, ] * unfished_default@params@w,
    kernel          = "default (beta=100, sigma=1.3)"
  ),
  data.frame(
    w               = unfished_highest_amp@params@w,
    biomass_density = unfished_highest_amp@n[last_highamp, 1, ] * unfished_highest_amp@params@w,
    kernel          = "highest-amplitude (beta=50, sigma=0.8)"
  )
)

ggplot(spectrum_df, aes(x = w, y = biomass_density, color = kernel)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Body weight (g)", y = "Biomass density (N x w)",
       title = "Size spectra at effort = 0",
       subtitle = "Dashed line = w_mat = 10, the default gear knife-edge selectivity cutoff") +
  theme_minimal()
```

Both curves drop steeply right at `w_mat = 10`, but they also cross a couple of times near the cutoff, so it's hard to tell by eye which kernel actually has more catchable biomass. Quantifying it directly (correctly, weighting by the log-spaced bin width `dw`, the same integral `getBiomass()` uses internally):

```{r}
#| message: false
#| warning: false
catchable_fraction <- function(sim, cutoff = 10) {
  last         <- dim(sim@n)[1]
  w            <- sim@params@w
  dw           <- sim@params@dw
  bm_density   <- sim@n[last, 1, ] * w * dw
  total_bm     <- sum(bm_density)
  catchable_bm <- sum(bm_density[w >= cutoff])
  data.frame(total_bm = total_bm, catchable_bm = catchable_bm,
             catchable_fraction = catchable_bm / total_bm)
}

catchable_summary <- rbind(
  data.frame(kernel = "default (beta=100, sigma=1.3)",
             catchable_fraction(unfished_default)),
  data.frame(kernel = "highest-amplitude (beta=50, sigma=0.8)",
             catchable_fraction(unfished_highest_amp))
)
knitr::kable(catchable_summary)
```

Confirmed. Despite having roughly 7× more total biomass, the highest-amplitude kernel has slightly less absolute catchable biomass than the default (`5.98e-6` vs `7.91e-6`), and its `catchable_fraction` is about 10× lower (`0.0033%` vs `0.031%`). Nearly all of that kernel's extra standing stock is juveniles that never cross the `w_mat = 10` selectivity cutoff.So the "more biomass" seen in the size-spectrum plot is largely fictitious from the fishery's perspective, and is the primary explanation for the yield gap.

### What This Shows

Feeding-kernel shape sets oscillation amplitude directly: narrower, more concentrated kernels sustain larger-amplitude cycles than the default. Fishing effort reduces biomass smoothly in both kernels without finding a maximum-sustainable-yield peak in the tested range, most likely because eased predation pressure lets the resource partially recover as the fish population is fished down. And a striking yield/biomass mismatch between kernels — more standing stock but far less catch — turns out to be a size-selectivity effect: most of the higher-amplitude kernel's extra biomass is juveniles below the catchable-size cutoff, not extra mature stock.

------------------------------------------------------------------------

## Why Are There So Many More Juveniles Than Adults?

Both size spectra above show a steep drop in biomass right at `w_mat = 10` ; several orders of magnitude fewer individuals ever reach a catchable, mature size than exist as juveniles. Going back to the original Day 18-style regime (`balance = TRUE`, `resource_decrease = r_d[1] = 0.001`, default perturbation, reusing the sim already computed in `results_balance_true[[1]]`), the natural first check is whether adults or juveniles oscillate more:

```{r}
#| message: false
#| warning: false
#| output: false
track_life_stages <- function(sim, w_mat = 10) {
  bm_juv   <- getBiomass(sim, max_w = w_mat)[, "Anchovy"]
  bm_adult <- getBiomass(sim, min_w = w_mat)[, "Anchovy"]
  t        <- as.numeric(names(bm_juv))
  data.frame(t = t, bm_juv = bm_juv, bm_adult = bm_adult)
}

day18_sim     <- results_balance_true[[1]]$sim
life_stage_bm <- track_life_stages(day18_sim)
mask          <- life_stage_bm$t > 550
```

```{r}
#| message: false
#| warning: false
plot(NULL, xlim = c(550, 600),
     ylim = range(c(life_stage_bm$bm_juv[mask], life_stage_bm$bm_adult[mask])),
     xlab = "Time (years)", ylab = "Biomass",
     main = "Juvenile vs adult biomass (balance = TRUE, r_d = 0.001)")
lines(life_stage_bm$t[mask], life_stage_bm$bm_juv[mask],   col = "darkgreen", lwd = 2)
lines(life_stage_bm$t[mask], life_stage_bm$bm_adult[mask], col = "purple",    lwd = 2)
legend("topright", legend = c("Juveniles (w < 10)", "Adults (w >= 10)"),
       col = c("darkgreen", "purple"), lwd = 2)
```

Juveniles and adults sit at very different biomass scales, so relative amplitude (`amp / mean`) is the fair comparison, not absolute amplitude:

```{r}
#| message: false
#| warning: false
rel_amp <- function(x) (max(x) - min(x)) / mean(x)

life_stage_summary <- data.frame(
  stage   = c("Juvenile (w < 10)", "Adult (w >= 10)"),
  mean_bm = c(mean(life_stage_bm$bm_juv[mask]),  mean(life_stage_bm$bm_adult[mask])),
  amp     = c(max(life_stage_bm$bm_juv[mask]) - min(life_stage_bm$bm_juv[mask]),
             max(life_stage_bm$bm_adult[mask]) - min(life_stage_bm$bm_adult[mask])),
  rel_amp = c(rel_amp(life_stage_bm$bm_juv[mask]), rel_amp(life_stage_bm$bm_adult[mask]))
)
knitr::kable(life_stage_summary)
```

The result is the opposite of expected. Adults show higher relative amplitude than juveniles (\~2.11 vs \~1.63) and proportionally, the rare mature fish swing more than the abundant juveniles do, not less. But that answers a different question from the one that actually matters here: how much either life stage oscillates around its own mean says nothing about why the standing-stock ratio between juveniles and adults is so lopsided in the first place. That's a demographic question, more about survivorship through growth, cumulative mortality, and how long an individual takes to grow from `w_min` to `w_mat` and not a dynamical one about oscillation amplitude.

------------------------------------------------------------------------

## Theoretical Grounding: Strogatz, Chapter 8

Alongside the coding, the next stretch of work is going through *Nonlinear Dynamics and Chaos* (Strogatz), sections 8.3–8.7, plus the accompanying exercises. These chapters line up almost exactly with the open questions this project has been circling since Day 18: what kind of bifurcation this is, whether the parked `beta = 150, sigma = 0.8` cell is chaos, quasiperiodicity, or a slow transient, and how to test for hysteresis properly rather than by eye.

------------------------------------------------------------------------

## What's Next

- **Plot maximum and minimum separately:** Instead of plotting max − min, to get a bifurcation-diagram look at the amplitude data rather than a single collapsed amplitude curve.

- **Try a different perturbation type:** To see whether the same convergence-to-attractor result holds no matter what form the perturbation takes, rather than relying solely on the mature-stock knock-down used so far.

- **Working through the Nonlinear Dynamics book:** Continuing Strogatz sections 8.3–8.7 on global bifurcations, hysteresis, quasiperiodicity, and Poincaré maps, to properly ground the bifurcation classification and diagnose the parked chaos-looking cell.

- **Think on why there is so much biomass accumulated at juvenile stages:** The survivorship-through-growth question raised above but not resolved.
