---
title: "Day 41: Oscillation Walkthrough"
author: "Samia"
date: "2026-08-17"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
categories: [research, mizer, anchovy, fishing, cannibalism, cultivation-effect, extinction, blog]
description: "Walk through of how I built the oscillations in this project."
---

Everything in this project's own research posts is produced through layers of helper functions accumulated over many days (`anchovy_params()`, `cannibalism_fishing_probe()`, `run_full_sweep_suite()`, and so on, all in `finding_jams/40_changed_mort_experiments.R`). None of that machinery is visible in those results, which makes the model itself hard to talk someone through. This post builds the same model -- `theta=0.3`, `interaction_resource=1`, `knife_edge=10`, the aim model behind this whole project -- once, directly, with nothing hidden behind a helper function. Every piece is exactly what the rest of this project uses, not a simplified stand-in for it: the paper's own parameters, the exact analytic resource stepper adopted on Day 40, the same settle-and-rescale procedure used to reach the oscillating state everywhere else.

```{r setup}
#| message: false
library(mizer)
library(ggplot2)
library(dplyr)
```

## A.1 The Parameters

These come from the paper this project is based on -- physiological rates (growth, reproduction), the predation kernel's shape, and the background resource's own dynamics. Nothing here is mizer-specific; it's just numbers.

```{r params}
p <- list(
  dt = 0.001, dx = 0.1, w_min = 0.0003, w_inf = 66.5,
  ppmr_min = 100, ppmr_max = 30000, gamma = 750, alpha = 0.85, K = 0.1,
  mu_l = 0, w_l = 0.03, rho_l = 5,
  mu_0 = 1, rho_b = -0.25,
  w_s = 0.5, rho_s = 1,
  w_mat = 10, rho_m = 15, rho_inf = 0.2, epsilon_R = 0.1,
  w_pp_cutoff = 0.1, r0 = 10, a0 = 100, i0 = 100, rho = 0.85, lambda = 2
)

theta                 <- 0.3   # fraction of baseline cannibalism strength kept
interaction_resource  <- 1     # this species' access to the background resource
knife_edge_size       <- 10    # fishing gear cutoff, in grams
```

## A.2 Two Small Functions mizer Needs

mizer looks up a predator's diet shape and the resource's own dynamics by name, so both need a matching function defined before `newMultispeciesParams()` is called.

**The predation kernel** says which prey sizes a predator will eat: anything within `[ppmr_min, ppmr_max]` of its own mass, in a flat "box" -- no size in that range is preferred over any other.

```{r kernel}
norm_box_pred_kernel <- function(ppmr, ppmr_min, ppmr_max) {
  phi <- rep(1, length(ppmr))
  phi[ppmr > ppmr_max | ppmr < ppmr_min] <- 0
  phi[1] <- 0
  phi / (sum(phi) * (log(ppmr[2]) - log(ppmr[1])))
}
```

**The resource's own dynamics** follow the logistic-plus-immigration equation $\dot n = \text{immigration} + (r - \mu)n - (r/K)n^2$: ordinary logistic growth toward a carrying capacity $K$, grazed down by predation mortality $\mu$ from the fish population, plus a constant trickle of immigration (production entering from outside the modelled size range, keeping the very smallest size classes topped up). mizer's own default resource stepper takes a single explicit-Euler step of this equation each `dt`; Day 40 of this project replaced that with the exact analytic solution of the same equation instead, adopted here unchanged because it's exact at any step size rather than an approximation that only holds for small `dt`.

```{r resource-dynamics}
exact_logistic_immigration_step <- function(n, rate, capacity, immigration,
                                            mortality, dt) {
  result <- n
  active <- is.finite(n) & is.finite(rate) & is.finite(capacity) &
    is.finite(immigration) & is.finite(mortality) &
    rate > 0 & capacity > 0 & immigration >= 0
  if (dt == 0 || !any(active)) return(result)

  n0 <- n[active]; r <- rate[active]; k <- capacity[active]
  i  <- immigration[active]; mu <- mortality[active]
  a  <- r - mu; b <- r / k
  next_n <- numeric(length(n0))

  # With immigration the quadratic has one positive and one negative root;
  # the alternative root formulae avoid cancellation when abs(a) is large.
  has_immigration <- i > 0
  if (any(has_immigration)) {
    idx <- which(has_immigration)
    ai <- a[idx]; bi <- b[idx]; ii <- i[idx]
    d  <- sqrt(ai^2 + 4 * bi * ii)
    n_plus  <- ifelse(ai >= 0, (ai + d) / (2 * bi), 2 * ii / (d - ai))
    n_minus <- ifelse(ai <= 0, (ai - d) / (2 * bi), -2 * ii / (ai + d))
    ratio <- ((n0[idx] - n_plus) / (n0[idx] - n_minus)) * exp(-d * dt)
    next_n[idx] <- (n_plus - ratio * n_minus) / (1 - ratio)
  }

  # The zero-immigration limit is handled separately, including rate == mortality.
  no_immigration <- !has_immigration
  if (any(no_immigration)) {
    idx <- which(no_immigration)
    az <- a[idx]; bz <- b[idx]; nz <- n0[idx]
    value <- numeric(length(idx))
    zero_rate <- az == 0
    value[zero_rate] <- nz[zero_rate] / (1 + bz[zero_rate] * nz[zero_rate] * dt)

    positive <- az > 0 & nz > 0
    phi <- -expm1(-az[positive] * dt) / az[positive]
    value[positive] <- nz[positive] / (exp(-az[positive] * dt) + bz[positive] * nz[positive] * phi)

    negative <- az < 0
    exp_adt <- exp(az[negative] * dt)
    phi <- expm1(az[negative] * dt) / az[negative]
    value[negative] <- nz[negative] * exp_adt / (1 + bz[negative] * nz[negative] * phi)
    next_n[idx] <- value
  }

  # Round-off can only create tiny negative values; the analytic solution is
  # non-negative for non-negative initial abundance and immigration.
  result[active] <- pmax(next_n, 0)
  result
}

plankton_logistic <- function(params, n, n_pp, n_other, rates, dt = 0.1, ...) {
  exact_logistic_immigration_step(
    n = n_pp, rate = params@rr_pp, capacity = params@cc_pp,
    immigration = immigration, mortality = rates$resource_mort, dt = dt
  )
}
```

## A.3 Building the Model

One species, described by the parameters above, dropped into `newMultispeciesParams()`.

```{r build}
kappa <- p$a0 * exp(-6.9 * (p$lambda - 1))

species_params_df <- data.frame(
  species = "Anchovy", w_min = p$w_min, w_mat = p$w_mat, m = p$rho_inf + 2/3,
  w_inf = p$w_inf, erepro = p$epsilon_R, alpha = p$K, ks = 0, gamma = p$gamma,
  q = p$alpha, ppmr_min = p$ppmr_min, ppmr_max = p$ppmr_max,
  pred_kernel_type = "norm_box", h = Inf, R_max = Inf, linecolour = "brown",
  stringsAsFactors = FALSE
)

params <- newMultispeciesParams(
  species_params_df, no_w = round(log(p$w_inf / p$w_min) / p$dx),
  lambda = p$lambda, kappa = kappa, w_pp_cutoff = p$w_pp_cutoff,
  resource_dynamics = "plankton_logistic"
)
resource_rate(params) <- p$r0 * w_full(params)^(p$rho - 1)
immigration            <- p$i0 * w_full(params)^(-p$lambda) * exp(-6.9 * (p$lambda - 1))
```

**Cannibalism and resource access.** `interaction_matrix` controls how much of a species' own predation pressure lands on itself; setting it to `theta` rather than 1 is literally what "cutting cannibalism to 30%" means in this model.

```{r interaction}
interaction_matrix(params) <- theta
species_params(params)$interaction_resource <- interaction_resource
```

**Fishing gear.** A knife-edge selectivity: nothing below `knife_edge_size` grams is caught, everything above it is, at a fixed catchability.

```{r gear}
gp                 <- gear_params(params)
gp$sel_func        <- "knife_edge"
gp$knife_edge_size <- knife_edge_size
gp$catchability    <- 1
gear_params(params) <- gp
```

**Background mortality.** This is the one place this model deliberately departs from mizer's own default: instead of a flat mortality rate at every size, the paper specifies a U-shaped curve -- high for larvae, a minimum around half a gram, rising again toward the adult size. `ext_mort<-()` is used rather than writing to `@mu_b` directly so this survives any later parameter change that would otherwise silently recompute it from mizer's own defaults (the bug Day 40 of this project found and fixed).

```{r mortality}
w    <- w(params)
mu_b <- rep(0, length(w))
mu_b[w <= p$w_s] <- (p$mu_0 * (w / p$w_min)^p$rho_b)[w < p$w_s]
mu_s <- min(mu_b[w <= p$w_s])
mu_b[w >= p$w_s] <- (mu_s * (w / p$w_s)^p$rho_s)[w >= p$w_s]
mu_b <- mu_b + p$mu_l / (1 + (w / p$w_l)^p$rho_l)

mort    <- ext_mort(params)
mort[]  <- mu_b
ext_mort(params) <- mort
```

The model is now fully specified. Everything below just runs it.

## A.4 Does It Oscillate?

The rest of this project reaches the oscillating state through a hand-tuned, two-stage procedure: settle for 10 years from an arbitrary starting spectrum, then rescale down by `1e7` (the settling transient overshoots by many orders of magnitude) before continuing. That procedure is specific to the default first-order scheme -- this section uses mizer's own `projectToSteady()` instead, which projects forward until the population's own oscillation amplitude stops changing by more than `amplitude_tol` between checks, with no hand-tuned settle time or rescale constant. It also switches the *time*-stepping method to `tr_bdf2`, mizer's second-order-in-time scheme -- not `second_order_w` (the *spatial* flux scheme, left at its default here), which is a narrower, different test from the full second-order combination this project's own Day 41 caveat found trouble with. Whether this still shows the same roughly five-to-six year cycle found everywhere else in this project, or whether `projectToSteady()`/`tr_bdf2` changes it, is exactly what running this checks.

```{r settle}
p$method <- "tr_bdf2"   # time-stepping method only -- second_order_w (the spatial
                        # flux scheme) is left at its own default, deliberately

sim_unfished <- projectToSteady(params, t_max = 150, t_per = 0.2, dt = p$dt,
                                method = p$method, return_sim = TRUE,
                                amplitude_tol = 0.0001)
```

```{r plot-oscillation}
#| fig-cap: "Unfished biomass under projectToSteady()'s own tr_bdf2 convergence run -- whatever length that took to satisfy amplitude_tol=0.0001, up to its own t_max=150 cap, not a fixed window chosen in advance."
#| message: false
plot(getBiomass(sim_unfished))
```

## A.5 Its Yield Curve

Rather than a fixed-length projection and an arbitrary post-transient window, run `projectToSteady()` again, once per fishing effort -- `finalParams(sim_unfished)` hands the model's own converged state straight back in as the next run's starting parameters, no manual `initialN<-()`/`initialNResource<-()` bookkeeping needed. `projectToSteady()` reports the cycle length it actually detected in `attr(sim, "convergence")$period`; averaging yield only over that final, detected period -- not a hand-picked number of years -- gives a genuinely smooth per-cycle mean rather than one that depends on exactly which stretch happened to be sampled.

```{r yield-curve}
effort_seq <- c(1, 2, 3, 5, 9, 20, 50, 100)
params <- finalParams(sim_unfished)

mean_yield <- sapply(effort_seq, function(effort) {
  sim <- projectToSteady(params, dt = p$dt, method = p$method, t_per = 0.2,
                 return_sim = TRUE, effort = effort)
  conv <- attr(sim, "convergence")
  start <- conv$years - conv$period
  mean(getYield(sim)[getTimes(sim) >= start, ])
})

yield_df <- data.frame(effort = effort_seq, yield = mean_yield)
```

```{r plot-yield}
#| fig-cap: "Yield vs. constant fishing effort for the oscillating anchovy model."
ggplot(yield_df, aes(effort, yield)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(x = "Fishing effort (Constant, log scale)", y = "Mean yield",
       title = "Yield curve: theta=0.3, interaction_resource=1, knife_edge=10") +
  theme_minimal()
```

Under the project's own default settle procedure and first-order scheme, this curve is **not** a clean single peak. This project's own finer sweeps (at this exact parameter point but with a different FMort) found a local peak around effort 9-12, a genuine dip through 15-20, and then a second, larger rise all the way out to 100. This was driven by a compositional shift, not a healthy fishery: sustained pressure on the harvestable size range relieves the cannibalism pressure on juveniles enough that they boom into the space it leaves behind, and that juvenile boom keeps throwing off catchable biomass even once the harvestable size classes are nearly wiped out. Whether `projectToSteady()`/`tr_bdf2`'s own different seed state reproduces that same shape, or something else, is an open question this rewrite is set up to answer rather than a foregone conclusion -- worth comparing directly against the earlier figure once this actually runs, not assuming the two must agree.
